---
title: "LFO-CV of Cherry Blossoms Data"
author: "Paul Bürkner & Aki Vehtari"
output:
  html_document:
    theme: default
encoding: UTF-8
editor_options: 
  chunk_output_type: console
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

---

# {.tabset}

## Setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.width = 10)
```

```{r general_options, cache=FALSE}
source("sim_functions.R")
library(tidyverse)
library(brms)
options(width = 160, mc.cores = 2)
theme_set(theme_default())
```

```{r}
cherry <- read.csv("data/cherry_blossoms.csv")
cherry_temp <- cherry[!is.na(cherry$temp), ]
cherry_doy <- cherry[!is.na(cherry$doy), ]
```

## Model

Fit a GP model:
```{r}
fit_cb <- brm(
  formula = bf(doy ~ gp(year, k = 20, c = 5/4)), 
  data = cherry_doy, 
  prior = prior(normal(0, 0.1), class = lscale, coef = gpyear),
  chain = 2, warmup = 4000, iter = 7000, inits = 0,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 5838234, file = "models/fit_cb"
)
plot(marginal_effects(fit_cb), points = TRUE)
```

## LFO-CV with 1-SAP {.tabset}

```{r}
L <- 100
M <- 1
```

### ELPD

Perform exact LFO-CV:
```{r}
cb_elpd_1sap_exact <- compute_lfo(
  fit_cb, type = "exact", M = M, L = L, 
  file = "results/cb_elpd_1sap_exact.rds"
)
summarize_elpds(cb_elpd_1sap_exact)
```

Perform approximate LFO-CV:
```{r}
cb_elpd_1sap_approx <- compute_lfo(
  fit_cb, type = "approx", M = M, L = L, k_thres = 0.6,
  file = "results/cb_elpd_1sap_approx.rds"
)
summarize_elpds(cb_elpd_1sap_approx)
```

Perform approximate LOO-CV:
```{r}
(loo_cb <- loo(fit_cb, newdata = cherry_doy[-seq_len(L), ]))
```

Analysis of convergence:
```{r}
conv_exact <- attributes(cb_elpd_1sap_exact)$conv
names(conv_exact) <- seq_along(conv_exact)
predicted <- !sapply(conv_exact, is.null)
conv_exact <- conv_exact[predicted]

conv_approx <- attributes(cb_elpd_1sap_approx)$conv
names(conv_approx) <- seq_along(conv_approx)
conv_approx <- conv_approx[predicted]

conv_1sap <- tibble(
  elpd_exact = na.omit(cb_elpd_1sap_exact),
  elpd_approx = na.omit(cb_elpd_1sap_approx)
) %>%
  mutate(
    i = L + 1:n(),
    elpd_diff = elpd_approx - elpd_exact,
    elpd_abs_diff = abs(elpd_diff),
    conv_exact = conv_exact,
    conv_approx = conv_approx
  ) %>%
  arrange(desc(elpd_abs_diff))

conv_1sap
```

### RMSE

Perform exact LFO-CV:
```{r}
# cb_rmse_1sap_exact <- compute_lfo(
#   fit_cb, type = "exact", M = M, L = L, criterion = "rmse",
#   file = "results/cb_rmse_1sap_exact.rds"
# )
# summarize_elpds(cb_rmse_1sap_exact)
```

Perform approximate LFO-CV:
```{r}
# cb_rmse_1sap_approx <- compute_lfo(
#   fit_cb, type = "approx", M = M, L = L, criterion = "rmse",
#   file = "results/cb_rmse_1sap_approx.rds"
# )
# summarize_elpds(cb_rmse_1sap_approx)
```


## LFO-CV with 4-SAP {.tabset}

```{r}
L <- 100
M <- 4
```

### ELPD

Perform exact LFO-CV:
```{r}
cb_elpd_4sap_exact <- compute_lfo(
  fit_cb, type = "exact", M = M, L = L, 
  file = "results/cb_elpd_4sap_exact.rds"
)
summarize_elpds(cb_elpd_4sap_exact) 
```

Perform approximate LFO-CV:
```{r}
cb_elpd_4sap_approx <- compute_lfo(
  fit_cb, type = "approx", M = M, L = L, k_thres = 0.6,
  file = "results/cb_elpd_4sap_approx.rds"
)
summarize_elpds(cb_elpd_4sap_approx)
```

Analysis of convergence:
```{r}
conv_exact <- attributes(cb_elpd_4sap_exact)$conv
names(conv_exact) <- seq_along(conv_exact)
predicted <- !sapply(conv_exact, is.null)
conv_exact <- conv_exact[predicted]

conv_approx <- attributes(cb_elpd_4sap_approx)$conv
names(conv_approx) <- seq_along(conv_approx)
conv_approx <- conv_approx[predicted]

conv_4sap <- tibble(
  elpd_exact = na.omit(cb_elpd_4sap_exact),
  elpd_approx = na.omit(cb_elpd_4sap_approx)
) %>%
  mutate(
    i = L + 1:n(),
    elpd_diff = elpd_approx - elpd_exact,
    elpd_abs_diff = abs(elpd_diff),
    conv_exact = conv_exact,
    conv_approx = conv_approx
  ) %>%
  arrange(desc(elpd_abs_diff))

conv_4sap
```

### RMSE

Perform exact LFO-CV:
```{r}
# cb_rmse_4sap_exact <- compute_lfo(
#   fit_cb, type = "exact", M = M, L = L, criterion = "rmse",
#   file = "results/cb_rmse_4sap_exact.rds"
# )
# summarize_elpds(cb_rmse_4sap_exact)
```

Perform approximate LFO-CV:
```{r}
# cb_rmse_4sap_approx <- compute_lfo(
#   fit_cb, type = "approx", M = M, L = L, criterion = "rmse",
#   file = "results/cb_rmse_4sap_approx.rds"
# )
# summarize_elpds(cb_rmse_4sap_approx)
```
