---
title: "LFO-CV of Cherry Blossoms Data"
author: "Paul Bürkner, Jonah Gabry, & Aki Vehtari"
output:
  html_document:
    theme: default
encoding: UTF-8
editor_options: 
  chunk_output_type: console
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

---

# {.tabset}

## Setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.width = 10)
```

```{r general_options, cache=FALSE}
source("sim_functions.R")
library(tidyverse)
library(brms)
options(width = 160, mc.cores = 2)
theme_set(theme_default())
```

```{r}
cherry <- read.csv("data/cherry_blossoms.csv")
cherry_temp <- cherry[!is.na(cherry$temp), ]
cherry_doy <- cherry[!is.na(cherry$doy), ]
```

## Model

Fit a GP model:
```{r}
fit_cb <- brm(
  formula = bf(doy ~ gp(year, k = 20, c = 5/4)), 
  data = cherry_doy, 
  prior = prior(normal(0, 0.1), class = lscale, coef = gpyear),
  chain = 2, warmup = 4000, iter = 7000, inits = 0,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 5838234, file = "models/fit_cb"
)
plot(marginal_effects(fit_cb), points = TRUE)
```

```{r}
N <- NROW(cherry_doy)
L <- 100
k_thres <- 0.7
```

## LFO-CV with 1-SAP {.tabset}

### ELPD

Perform exact LFO-CV:
```{r}
cb_elpd_1sap_exact <- compute_lfo(
  fit_cb, type = "exact", M = 1, L = L, factorize = TRUE,
  file = "results/cb_elpd_1sap_exact.rds"
)
```

Perform approximate LFO-CV (forward):
```{r}
cb_elpd_1sap_approx_fw <- compute_lfo(
  fit_cb, type = "approx", M = 1, L = L, 
  k_thres = k_thres, mode = "forward", factorize = TRUE,
  file = "results/cb_elpd_1sap_approx_fw.rds"
)
```

Perform approximate LFO-CV (backward):
```{r}
cb_elpd_1sap_approx_bw <- compute_lfo(
  fit_cb, type = "approx", M = 1, L = L, 
  k_thres = k_thres, mode = "backward", factorize = TRUE,
  file = "results/cb_elpd_1sap_approx_bw.rds"
)
```

Perform approximate LFO-CV (combined):
```{r}
cb_elpd_1sap_approx_cb <- compute_lfo(
  fit_cb, type = "approx", M = 1, L = L, 
  k_thres = k_thres, mode = "combined", factorize = TRUE,
  file = "results/cb_elpd_1sap_approx_cb.rds"
)
```

```{r}
summarize_elpds(cb_elpd_1sap_exact)
summarize_elpds(cb_elpd_1sap_approx_fw)
summarize_elpds(cb_elpd_1sap_approx_bw)
summarize_elpds(cb_elpd_1sap_approx_cb)
```

Perform approximate LOO-CV:
```{r}
(loo_cb <- loo(fit_cb, newdata = cherry_doy[-seq_len(L), ]))
```

Analysis of convergence and pointwise accuracy:
```{r}
conv_exact <- attributes(cb_elpd_1sap_exact)$conv
names(conv_exact) <- seq_along(conv_exact)
predicted <- !sapply(conv_exact, is.null)
conv_exact <- conv_exact[predicted]

conv_approx_fw <- attributes(cb_elpd_1sap_approx_fw)$conv
names(conv_approx_fw) <- seq_along(conv_approx_fw)
conv_approx_fw <- conv_approx_fw[predicted]

conv_approx_bw <- attributes(cb_elpd_1sap_approx_bw)$conv
names(conv_approx_bw) <- seq_along(conv_approx_bw)
conv_approx_bw <- conv_approx_bw[predicted]

conv_approx_cb <- attributes(cb_elpd_1sap_approx_cb)$conv
names(conv_approx_cb) <- seq_along(conv_approx_cb)
conv_approx_cb <- conv_approx_cb[predicted]

cb_pw_elpd_1sap <- tibble(
  elpd_exact = na.omit(cb_elpd_1sap_exact),
  elpd_approx_fw = na.omit(cb_elpd_1sap_approx_fw),
  elpd_approx_bw = na.omit(cb_elpd_1sap_approx_bw),
  elpd_approx_cb = na.omit(cb_elpd_1sap_approx_cb),
) %>%
  mutate(
    i = L + 1:n(),
    elpd_diff_fw = elpd_approx_fw - elpd_exact,
    elpd_abs_diff_fw = abs(elpd_diff_fw),
    k_fw = na.omit(attributes(cb_elpd_1sap_approx_fw)$ks),
    reff_fw = na.omit(attributes(cb_elpd_1sap_approx_fw)$reffs),
    elpd_diff_bw = elpd_approx_bw - elpd_exact,
    elpd_abs_diff_bw = abs(elpd_diff_bw),
    k_bw = na.omit(attributes(cb_elpd_1sap_approx_bw)$ks),
    reff_bw = na.omit(attributes(cb_elpd_1sap_approx_bw)$reffs),
    elpd_diff_cb = elpd_approx_cb - elpd_exact,
    elpd_abs_diff_cb = abs(elpd_diff_cb),
    k_cb = na.omit(attributes(cb_elpd_1sap_approx_cb)$ks),
    reff_cb = na.omit(attributes(cb_elpd_1sap_approx_cb)$reffs),
    conv_exact = conv_exact,
    conv_approx_fw = conv_approx_fw,
    conv_approx_bw = conv_approx_bw,
    conv_approx_cb = conv_approx_cb
  ) %>%
  arrange(desc(elpd_abs_diff_cb))
```

Plot khats:
```{r}
plot_ks(cb_pw_elpd_1sap$k_fw, cb_pw_elpd_1sap$i, k_thres = k_thres)
plot_ks(cb_pw_elpd_1sap$k_bw, cb_pw_elpd_1sap$i, k_thres = k_thres)
plot_ks(cb_pw_elpd_1sap$k_cb, cb_pw_elpd_1sap$i, k_thres = k_thres)
```

Plot relative efficiencies:
```{r}
cb_pw_elpd_1sap %>% 
  gather("mode", "reff", starts_with("reff")) %>%
  ggplot(aes(i, reff)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_point()
```

Plot pointwise accuracies:
```{r}
cb_pw_elpd_1sap %>% 
  gather("mode", "elpd_approx", starts_with("elpd_approx")) %>%
  ggplot(aes(elpd_exact, elpd_approx)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate ELPD", x = "Exact ELPD") + 
  theme_bw()
```


## LFO-CV with 4-SAP {.tabset}

### ELPD

Perform exact LFO-CV:
```{r}
cb_elpd_4sap_exact <- compute_lfo(
  fit_cb, type = "exact", M = 4, L = L, factorize = TRUE,
  file = "results/cb_elpd_4sap_exact.rds"
)
```

Perform approximate LFO-CV (forward):
```{r}
cb_elpd_4sap_approx_fw <- compute_lfo(
  fit_cb, type = "approx", M = 4, L = L, 
  k_thres = k_thres, mode = "forward", factorize = TRUE,
  file = "results/cb_elpd_4sap_approx_fw.rds"
)
```

Perform approximate LFO-CV (backward):
```{r}
cb_elpd_4sap_approx_bw <- compute_lfo(
  fit_cb, type = "approx", M = 4, L = L, 
  k_thres = k_thres, mode = "backward", factorize = TRUE,
  file = "results/cb_elpd_4sap_approx_bw.rds"
)
```

Perform approximate LFO-CV (combined):
```{r}
cb_elpd_4sap_approx_cb <- compute_lfo(
  fit_cb, type = "approx", M = 4, L = L, 
  k_thres = k_thres, mode = "combined", factorize = TRUE,
  file = "results/cb_elpd_4sap_approx_cb.rds"
)
```

```{r}
summarize_elpds(cb_elpd_4sap_exact)
summarize_elpds(cb_elpd_4sap_approx_fw)
summarize_elpds(cb_elpd_4sap_approx_bw)
summarize_elpds(cb_elpd_4sap_approx_cb)
```

Analysis of convergence:
```{r}
conv_exact <- attributes(cb_elpd_4sap_exact)$conv
names(conv_exact) <- seq_along(conv_exact)
predicted <- !sapply(conv_exact, is.null)
conv_exact <- conv_exact[predicted]

conv_approx_fw <- attributes(cb_elpd_4sap_approx_fw)$conv
names(conv_approx_fw) <- seq_along(conv_approx_fw)
conv_approx_fw <- conv_approx_fw[predicted]

conv_approx_bw <- attributes(cb_elpd_4sap_approx_bw)$conv
names(conv_approx_bw) <- seq_along(conv_approx_bw)
conv_approx_bw <- conv_approx_bw[predicted]

conv_approx_cb <- attributes(cb_elpd_4sap_approx_cb)$conv
names(conv_approx_cb) <- seq_along(conv_approx_cb)
conv_approx_cb <- conv_approx_cb[predicted]

cb_pw_elpd_4sap <- tibble(
  elpd_exact = na.omit(cb_elpd_4sap_exact),
  elpd_approx_fw = na.omit(cb_elpd_4sap_approx_fw),
  elpd_approx_bw = na.omit(cb_elpd_4sap_approx_bw),
  elpd_approx_cb = na.omit(cb_elpd_4sap_approx_cb)
) %>%
  mutate(
    i = L + 1:n(),
    elpd_diff_fw = elpd_approx_fw - elpd_exact,
    elpd_abs_diff_fw = abs(elpd_diff_fw),
    k_fw = na.omit(attributes(cb_elpd_4sap_approx_fw)$ks),
    reff_fw = na.omit(attributes(cb_elpd_4sap_approx_fw)$reffs),
    elpd_diff_bw = elpd_approx_bw - elpd_exact,
    elpd_abs_diff_bw = abs(elpd_diff_bw),
    k_bw = na.omit(attributes(cb_elpd_4sap_approx_bw)$ks),
    reff_bw = na.omit(attributes(cb_elpd_4sap_approx_bw)$reffs),
    elpd_diff_cb = elpd_approx_cb - elpd_exact,
    elpd_abs_diff_cb = abs(elpd_diff_cb),
    k_cb = na.omit(attributes(cb_elpd_4sap_approx_cb)$ks),
    reff_cb = na.omit(attributes(cb_elpd_4sap_approx_cb)$reffs),
    conv_exact = conv_exact,
    conv_approx_fw = conv_approx_fw,
    conv_approx_bw = conv_approx_bw,
    conv_approx_cb = conv_approx_cb
  ) %>%
  arrange(desc(elpd_abs_diff_cb))
```

Plot khats:
```{r}
plot_ks(cb_pw_elpd_4sap$k_fw, cb_pw_elpd_4sap$i, k_thres = k_thres)
plot_ks(cb_pw_elpd_4sap$k_bw, cb_pw_elpd_4sap$i, k_thres = k_thres)
plot_ks(cb_pw_elpd_4sap$k_cb, cb_pw_elpd_4sap$i, k_thres = k_thres)
```

Plot relative efficiencies:
```{r}
cb_pw_elpd_4sap %>% 
  gather("mode", "reff", starts_with("reff")) %>%
  ggplot(aes(i, reff)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_point()
```

Plot pointwise accuracies:
```{r}
cb_pw_elpd_4sap %>% 
  gather("mode", "elpd_approx", starts_with("elpd_approx")) %>%
  ggplot(aes(elpd_exact, elpd_approx)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate ELPD", x = "Exact ELPD") + 
  theme_bw()
```

