---
title: "LFO-CV of Lake Huron Data"
author: "Paul Bürkner, Jonah Gabry, & Aki Vehtari"
output:
  html_document:
    theme: default
encoding: UTF-8
editor_options: 
  chunk_output_type: console
---

<style>
pre {
  overflow-x: auto;
}
pre code {
  word-wrap: normal;
  white-space: pre;
}
</style>

---

# {.tabset}

## Setup

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.width = 10)
```

```{r general_options, cache=FALSE}
source("sim_functions.R")
library(tidyverse)
library(brms)
options(width = 160, mc.cores = 2)
theme_set(theme_default())
```

```{r}
data("LakeHuron")
N <- length(LakeHuron)
df <- data.frame(
  y = as.numeric(LakeHuron),
  year = as.numeric(time(LakeHuron)),
  time = 1:N
)
```

## Model

Fit an AR(4) model:
```{r}
fit_lh <- brm(
  y | mi() ~ 1, 
  data = df, 
  autocor = cor_ar(~time, p = 4), 
  prior = prior(normal(0, 0.5), class = "ar"),
  chains = 2, warmup = 1000, iter = 4000,
  control = list(adapt_delta = 0.99),
  seed = 5838296, file = "models/fit_lh"
)
summary(fit_lh)
```

```{r}
preds <- posterior_predict(fit_lh)
preds <- cbind(
  Estimate = colMeans(preds), 
  Q5 = apply(preds, 2, quantile, probs = 0.05),
  Q95 = apply(preds, 2, quantile, probs = 0.95)
)

ggplot(cbind(df, preds), aes(x = year, y = Estimate)) +
  geom_smooth(aes(ymin = Q5, ymax = Q95), stat = "identity", size = 0.5) +
  geom_point(aes(y = y)) + 
  labs(y = "Water Level (ft)", x = "Year")
```

```{r}
L <- 20
k_thres <- 0.7
```

## LFO-CV of 1-SAP {.tabset}

### ELPD

```{r, results="hide", warning=FALSE, message=FALSE}
lh_elpd_1sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = 1, L = L,
  file = "results/lh_elpd_1sap_exact.rds"
)
lh_elpd_1sap_approx_fw <- compute_lfo(
  fit_lh, type = "approx", M = 1, L = L, 
  k_thres = k_thres, mode = "forward",
  file = "results/lh_elpd_1sap_approx_fw.rds"
)
lh_elpd_1sap_approx_bw <- compute_lfo(
  fit_lh, type = "approx", M = 1, L = L, 
  k_thres = k_thres, mode = "backward",
  file = "results/lh_elpd_1sap_approx_bw.rds"
)
lh_elpd_1sap_approx_cb <- compute_lfo(
  fit_lh, type = "approx", M = 1, L = L, 
  k_thres = k_thres, mode = "combined",
  file = "results/lh_elpd_1sap_approx_cb.rds"
)
```

```{r}
summarize_elpds(lh_elpd_1sap_exact)
summarize_elpds(lh_elpd_1sap_approx_fw)
summarize_elpds(lh_elpd_1sap_approx_bw)
summarize_elpds(lh_elpd_1sap_approx_cb)
```

Perform approximate LOO-CV:
```{r}
(loo_lh <- loo(log_lik(fit_lh)[, (L + 1):N]))
```

Analysis of convergence and pointwise accuracy:
```{r}
lh_pw_elpd_1sap <- tibble(
  i = L:(N - 1),
  elpd_exact = na.omit(lh_elpd_1sap_exact),
  elpd_approx_fw = na.omit(lh_elpd_1sap_approx_fw),
  k_fw = na.omit(attributes(lh_elpd_1sap_approx_fw)$ks),
  reff_fw = na.omit(attributes(lh_elpd_1sap_approx_fw)$reffs),
  elpd_approx_bw = na.omit(lh_elpd_1sap_approx_bw),
  k_bw = na.omit(attributes(lh_elpd_1sap_approx_bw)$ks),
  reff_bw = na.omit(attributes(lh_elpd_1sap_approx_bw)$reffs),
  elpd_approx_cb = na.omit(lh_elpd_1sap_approx_cb),
  k_cb = na.omit(attributes(lh_elpd_1sap_approx_cb)$ks),
  reff_cb = na.omit(attributes(lh_elpd_1sap_approx_cb)$reffs)
)
```

Plot khats:
```{r}
plot_ks(lh_pw_elpd_1sap$k_fw, lh_pw_elpd_1sap$i, k_thres = k_thres)
plot_ks(lh_pw_elpd_1sap$k_bw, lh_pw_elpd_1sap$i, k_thres = k_thres)
plot_ks(lh_pw_elpd_1sap$k_cb, lh_pw_elpd_1sap$i, k_thres = k_thres)
```

Plot relative efficiencies:
```{r}
lh_pw_elpd_1sap %>% 
  gather("mode", "reff", starts_with("reff")) %>%
  ggplot(aes(i, reff)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_point()
```

Plot pointwise accuracies:
```{r}
lh_pw_elpd_1sap %>% 
  gather("mode", "elpd_approx", starts_with("elpd_approx")) %>%
  ggplot(aes(elpd_exact, elpd_approx)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate ELPD", x = "Exact ELPD") + 
  theme_bw()
```

### RMSE

```{r, results="hide", warning=FALSE, message=FALSE}
lh_rmse_1sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = 1, L = L, 
  criterion = "rmse", 
  file = "results/lh_rmse_1sap_exact.rds"
)
lh_rmse_1sap_approx_fw <- compute_lfo(
  fit_lh, type = "approx", M = 1, L = L, 
  criterion = "rmse", k_thres = k_thres, mode = "forward",
  file = "results/lh_rmse_1sap_approx_fw.rds"
)
lh_rmse_1sap_approx_bw <- compute_lfo(
  fit_lh, type = "approx", M = 1, L = L, 
  criterion = "rmse", k_thres = k_thres, mode = "backward",
  file = "results/lh_rmse_1sap_approx_bw.rds"
)
lh_rmse_1sap_approx_cb <- compute_lfo(
  fit_lh, type = "approx", M = 1, L = L, 
  criterion = "rmse", k_thres = k_thres, mode = "combined",
  file = "results/lh_rmse_1sap_approx_cb.rds"
)
```

```{r}
summarize_elpds(lh_rmse_1sap_exact)
summarize_elpds(lh_rmse_1sap_approx_fw)
summarize_elpds(lh_rmse_1sap_approx_bw)
summarize_elpds(lh_rmse_1sap_approx_cb)
```

Analysis of pointwise accuracy:
```{r}
lh_pw_rmse_1sap <- tibble(
  i = L:(N - 1),
  rmse_exact = na.omit(lh_rmse_1sap_exact),
  rmse_approx_fw = na.omit(lh_rmse_1sap_approx_fw),
  k_fw = na.omit(attributes(lh_rmse_1sap_approx_fw)$ks),
  reff_fw = na.omit(attributes(lh_rmse_1sap_approx_fw)$reffs),
  rmse_approx_bw = na.omit(lh_rmse_1sap_approx_bw),
  k_bw = na.omit(attributes(lh_rmse_1sap_approx_bw)$ks),
  reff_bw = na.omit(attributes(lh_rmse_1sap_approx_bw)$reffs),
  rmse_approx_cb = na.omit(lh_rmse_1sap_approx_cb),
  k_cb = na.omit(attributes(lh_rmse_1sap_approx_cb)$ks),
  reff_cb = na.omit(attributes(lh_rmse_1sap_approx_cb)$reffs)
)
```

Plot pointwise accuracies:
```{r}
lh_pw_rmse_1sap %>% 
  gather("mode", "rmse_approx", starts_with("rmse_approx")) %>%
  ggplot(aes(rmse_exact, rmse_approx)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate RMSE", x = "Exact RMSE") + 
  theme_bw()
```



## LFO-CV of 4-SAP {.tabset}

### ELPD

```{r, results="hide", warning=FALSE, message=FALSE}
lh_elpd_4sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = 4, L = L,
  file = "results/lh_elpd_4sap_exact.rds"
)
lh_elpd_4sap_approx_fw <- compute_lfo(
  fit_lh, type = "approx", M = 4, L = L,
  k_thres = k_thres, mode = "forward",
  file = "results/lh_elpd_4sap_approx_fw.rds"
)
lh_elpd_4sap_approx_bw <- compute_lfo(
  fit_lh, type = "approx", M = 4, L = L, 
  k_thres = k_thres, mode = "backward",
  file = "results/lh_elpd_4sap_approx_bw.rds"
)
lh_elpd_4sap_approx_cb <- compute_lfo(
  fit_lh, type = "approx", M = 4, L = L, 
  k_thres = k_thres, mode = "combined",
  file = "results/lh_elpd_4sap_approx_cb.rds"
)
```

```{r}
summarize_elpds(lh_elpd_4sap_exact)
summarize_elpds(lh_elpd_4sap_approx_fw)
summarize_elpds(lh_elpd_4sap_approx_bw)
summarize_elpds(lh_elpd_4sap_approx_cb)
```

Analysis of pointwise accuracy:
```{r}
lh_pw_elpd_4sap <- tibble(
  i = L:(N - 4),
  elpd_exact = na.omit(lh_elpd_4sap_exact),
  elpd_approx_fw = na.omit(lh_elpd_4sap_approx_fw),
  k_fw = na.omit(attributes(lh_elpd_4sap_approx_fw)$ks),
  reff_fw = na.omit(attributes(lh_elpd_4sap_approx_fw)$reffs),
  elpd_approx_bw = na.omit(lh_elpd_4sap_approx_bw),
  k_bw = na.omit(attributes(lh_elpd_4sap_approx_bw)$ks),
  reff_bw = na.omit(attributes(lh_elpd_4sap_approx_bw)$reffs),
  elpd_approx_cb = na.omit(lh_elpd_4sap_approx_cb),
  k_cb = na.omit(attributes(lh_elpd_4sap_approx_cb)$ks),
  reff_cb = na.omit(attributes(lh_elpd_4sap_approx_cb)$reffs)
)
```

Plot khats:
```{r}
plot_ks(lh_pw_elpd_4sap$k_fw, lh_pw_elpd_4sap$i, k_thres = k_thres)
plot_ks(lh_pw_elpd_4sap$k_bw, lh_pw_elpd_4sap$i, k_thres = k_thres)
plot_ks(lh_pw_elpd_4sap$k_cb, lh_pw_elpd_4sap$i, k_thres = k_thres)
```

Plot relative efficiencies:
```{r}
lh_pw_elpd_4sap %>% 
  gather("mode", "reff", starts_with("reff")) %>%
  ggplot(aes(i, reff)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_point()
```

Plot pointwise accuracies:
```{r}
lh_pw_elpd_4sap %>% 
  gather("mode", "elpd_approx", starts_with("elpd_approx")) %>%
  ggplot(aes(elpd_exact, elpd_approx)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate ELPD", x = "Exact ELPD") + 
  theme_bw()
```


### RMSE

```{r, results="hide", warning=FALSE, message=FALSE}
lh_rmse_4sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = 4, L = L, 
  criterion = "rmse",
  file = "results/lh_rmse_4sap_exact.rds"
)
lh_rmse_4sap_approx_fw <- compute_lfo(
  fit_lh, type = "approx", M = 4, L = L, 
  criterion = "rmse", k_thres = k_thres, mode = "forward",
  file = "results/lh_rmse_4sap_approx_fw.rds"
)
lh_rmse_4sap_approx_bw <- compute_lfo(
  fit_lh, type = "approx", M = 4, L = L, 
  criterion = "rmse", k_thres = k_thres, mode = "backward",
  file = "results/lh_rmse_4sap_approx_bw.rds"
)
lh_rmse_4sap_approx_cb <- compute_lfo(
  fit_lh, type = "approx", M = 4, L = L, 
  criterion = "rmse", k_thres = k_thres, mode = "combined",
  file = "results/lh_rmse_4sap_approx_cb.rds"
)
```

```{r}
summarize_elpds(lh_rmse_4sap_exact)
summarize_elpds(lh_rmse_4sap_approx_fw)
summarize_elpds(lh_rmse_4sap_approx_bw)
summarize_elpds(lh_rmse_4sap_approx_cb)
```

Analysis of pointwise accuracy:
```{r}
lh_pw_rmse_4sap <- tibble(
  i = L:(N - 4),
  rmse_exact = na.omit(lh_rmse_4sap_exact),
  rmse_approx_fw = na.omit(lh_rmse_4sap_approx_fw),
  k_fw = na.omit(attributes(lh_rmse_4sap_approx_fw)$ks),
  reff_fw = na.omit(attributes(lh_rmse_4sap_approx_fw)$reffs),
  rmse_approx_bw = na.omit(lh_rmse_4sap_approx_bw),
  k_bw = na.omit(attributes(lh_rmse_4sap_approx_bw)$ks),
  reff_bw = na.omit(attributes(lh_rmse_4sap_approx_bw)$reffs),
  rmse_approx_cb = na.omit(lh_rmse_4sap_approx_cb),
  k_cb = na.omit(attributes(lh_rmse_4sap_approx_cb)$ks),
  reff_cb = na.omit(attributes(lh_rmse_4sap_approx_cb)$reffs)
)
```

Plot pointwise accuracies:
```{r}
lh_pw_rmse_4sap %>% 
  gather("mode", "rmse_approx", starts_with("rmse_approx")) %>%
  ggplot(aes(rmse_exact, rmse_approx)) +
  facet_wrap(facets = "mode", nrow = 1, ncol = 3, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate RMSE", x = "Exact RMSE") + 
  theme_bw()
```
