---
title: "Approximate leave-future-out cross-validation for Bayesian time series models"
author: "Paul-Christian Bürkner $^{1*}$, Jonah Gabry $^2$, & Aki Vehtari $^1$"
date: |
  $^1$ Department of Computer Science, Aalto University, Finland\break
  $^2$ Applied Statistics Center and ISERP, Columbia University, USA \break
  $^*$ Corresponding author, Email: paul.buerkner@gmail.com
abstract: |
  One of the common goals of time series analysis is to use the observed series
  to inform predictions for future observations. In the absence of any actual
  new data to predict, cross-validation can be used to estimate a model's future
  predictive accuracy, for instance, for the purpose of model comparison or
  selection. Exact cross-validation for Bayesian models is often
  computationally expensive, but approximate cross-validation methods have been
  developed, most notably methods for leave-one-out cross-validation (LOO-CV).
  If the actual prediction task is to predict the future given the past, LOO-CV
  provides an overly optimistic estimate because the information from future
  observations is available to influence predictions of the past. To properly 
  account for the time series structure, we can use
  leave-future-out cross-validation (LFO-CV). Like exact LOO-CV, exact LFO-CV
  requires refitting the model many times to different subsets of the data.
  Using Pareto smoothed importance sampling, we propose a method for
  approximating exact LFO-CV that drastically reduces the computational costs
  while also providing informative diagnostics about the quality of the
  approximation.\linebreak\linebreak 
  Keywords: Time Series Analysis, Cross-Validation, Bayesian Inference, Pareto Smoothed Importance Sampling
lang: en-US
class: man
# figsintext: true
numbersections: true
encoding: UTF-8
bibliography: LFO-CV
biblio-style: apalike
output:
  bookdown::pdf_document2:
     citation_package: natbib
     keep_tex: true
     toc: false
header-includes:
   - \usepackage{amsmath}
   - \usepackage[utf8]{inputenc}
   - \usepackage[T1]{fontenc}
   - \usepackage{setspace}
   - \onehalfspacing
   - \setcitestyle{round}
   - \newcommand\numberthis{\addtocounter{equation}{1}\tag{\theequation}}
editor_options: 
  chunk_output_type: console
---

```{r setup, cache = FALSE, include = FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
options(knitr.kable.NA = '')
```

```{r packages, cache = FALSE, include = FALSE}
library(knitr)
library(kableExtra)
library(latex2exp)
library(tidyverse)
library(brms)
library(loo)
source("sim_functions.R")

# set ggplot theme
theme_set(bayesplot::theme_default())

# set rstan options
rstan::rstan_options(auto_write = TRUE)
options(mc.cores = max(1, parallel::detectCores() - 1))
```

```{r functions}
fmt <- function(x, digits = 1, ...) {
  format(x, digits = digits, nsmall = digits, ...)
}
```

# Introduction

A wide range of statistical models for time series have been developed, finding
applications in industry and nearly all empirical sciences 
[e.g., see @brockwell2002; @hamilton1994].
One common goal of a time series analysis is to use the observed
series to inform predictions for future time points. In this paper we will
assume a Bayesian approach to time series modeling, in which case if it is
possible to sample from the posterior _predictive_ distribution implied by a
given time series model, then it is straightforward to generate predictions as
far into the future as we want. When working in discrete time we will refer to
the task of predicting a sequence of $M$ future observations as $M$-step-ahead
prediction ($M$-SAP).

It is easy to evaluate the $M$-SAP performance of a time series
model by comparing the predictions to the observed sequence of $M$ future data
points once they become available. However, we would often like to estimate 
the future predictive performance of a model _before_ we are able to collect 
additional observations. If there are many competing models we may also need to
first decide which model (or which combination of the models) to
rely on for prediction [@geisser1979; @hoeting1999; @vehtari2002; @ando2010; 
@vehtari2012]. 

In the absence of new data, one general approach for evaluating a model's
predictive accuracy is cross-validation. The data is first split into two
subsets, then we fit the statistical model to the first subset and evaluate
predictive performance with the second subset. We may do this once or many
times, each time leaving out a different subset.

If the data points are not ordered in time, or if the goal is to assess the
non-time-dependent part of the model, then we can use leave-one-out
cross-validation (LOO-CV). For a data set with $N$ observations, we refit the
model $N$ times, each time leaving out one of the $N$ observations and assessing
how well the model predicts the left-out observation. Due to the number of
required refits, exact LOO-CV is computationally expensive, in particular when
performing full Bayesian inference and refitting the model means estimating a
new posterior distribution rather than a point estimate. But it is possible to
approximate exact LOO-CV using Pareto smoothed importance sampling 
[PSIS; @vehtari2017loo; @vehtari2019psis]. 
PSIS-LOO-CV only requires a single fit of the full model and has sensitive
diagnostics for assessing the validity of the approximation.

However, using LOO-CV with times series models is problematic if the goal is to
estimate the predictive performance for future time points. Leaving out only one
observation at a time will allow information from the future to influence
predictions of the past (i.e., data from times $t+1, t+2, \ldots,$ would inform
predictions for time $t$). Instead, to apply the idea of cross-validation to the
$M$-SAP case we can use what we will refer to as leave-*future*-out
cross-validation (LFO-CV). LFO-CV does not refer to one particular prediction
task but rather to various possible cross-validation approaches that all involve
some form of prediction of future time points.

Like exact LOO-CV, exact LFO-CV requires refitting the model many times to
different subsets of the data, which is computationally expensive, in particular
for full Bayesian inference. In this paper, we extend the ideas from PSIS-LOO-CV
and present PSIS-LFO-CV, an algorithm that typically only requires refitting a
time series model a small number times. This will make LFO-CV tractable for many
more realistic applications than previously possible, including time series
model averaging using stacking of predictive distributions [@yao2018].

The efficiency of PSIS-LFO-CV compared to exact LFO-CV relies on the
ability to compute samples from the posterior predictive distribution (required
for the importance sampling) in much less time than it takes to fully refit the
model. This assumption is most likely justified when estimating a model using
full Bayesian inference via MCMC, variational inference, or related methods as
they are very computationally intensive. We do not make any
assumptions about *how* samples from the posterior or the posterior predictive
density at a given point in time have been obtained.

Our proposed algorithm was motivated by the practical need for efficient
cross-validation tools for Bayesian time series models fit using generic
probabilistic programming frameworks, such as Stan [@carpenter2017], JAGS
[@jags], PyMC3 [@pymc3] and Pyro [@pyro]. These frameworks have become very
popular in recent years also for analysis of time series models. For many
models, inference could also be performed using sequential Monte Carlo (SMC)
[e.g.,@doucet2000; @andrieu2010] using, for example, the SMC-specific framework
Birch [@birch]. The implementation details of LFO-CV for SMC algorithms are
beyond the scope of this paper.\footnote{Most SMC algorithms use importance
sampling and LFO-CV could be obtained as a by-product, with computation
resembling the approach we present here. The proposal distribution at each step
and the applied "refit" approach (when the importance sampling weights become
degenerate) are specific to each SMC algorithm.}

The structure of the paper is as follows. In Section \@ref(m-sap), we introduce
the idea and various forms of $M$-step-ahead prediction and how to approximate
them using PSIS. In Section \@ref(simulations), we evaluate the accuracy of the
approximation using extensive simulations. Then, in Section \@ref(case-studies),
we provide two case studies demonstrating the application of PSIS-LFO-CV methods
to real data sets. In the first we model changes in the water level of Lake
Huron and in the second the date of the yearly cherry blossom in Kyoto. We end
in Section \@ref(discussion) with a discussion of the usefulness and limitations
of our approach.

# $M$-step-ahead predictions {#m-sap}

Assume we have a time series of observations $y = (y_1, y_2, \ldots, y_N)$ 
and let $L$ be the _minimum_ number of observations from the series that
we will require before making predictions for future data. Depending on the
application and how informative the data are, it may not be possible to make
reasonable predictions for $y_{i+1}$ based on $(y_1, \dots, y_{i})$ until $i$ is
large enough so that we can learn enough about the time series to predict future
observations. Setting $L=10$, for example, means that we will only assess
predictive performance starting with observation $y_{11}$, so that we
always have at least 10 previous observations to condition on.

In order to assess $M$-SAP performance we would like to compute the 
predictive densities

\begin{equation}
p(y_{i+1:M} \,|\, y_{1:i}) = 
  p(y_{i+1}, \ldots, y_{i + M} \,|\, y_{1},...,y_{i}) 
\end{equation}

for each $i \in \{L, \ldots, N - M\}$, where we use 
$y_{i+1:M} = (y_{i+1}, \ldots, y_{i + M})$ and $y_{1:i} = (y_{1}, \ldots, y_{i})$ 
to shorten the notation. As a global measure of predictive accuracy, we
can use the expected log predictive density [ELPD; @vehtari2017loo], which, 
for M-SAP, can be defined as

\begin{equation}
\label{ELPD}
{\rm ELPD} = \sum_{i=L}^{N - M} 
  \int p_t(\tilde{y}_{i+1:M}) \log p(\tilde{y}_{i+1:M} \,|\, y_{1:i})
  \, {\rm d} \, \tilde{y}_{i+1:M}.
\end{equation}

The distribution $p_t(\tilde{y}_{i+1:M})$ describes the true data generating
process for new data $\tilde{y}_{i+1:M}$. As these true data generating processes
are unknown, we approximate the ELPD using LFO-CV of the observed responses 
$y_{i+1:M}$, which constitute a particular realization of $\tilde{y}_{i+1:M}$.
Plugging in the realization $y_{i+1:M}$ for $\tilde{y}_{i+1:M}$ leads
to [c.f., @bernardo1994; @vehtari2012]:
 
\begin{equation}
{\rm ELPD}_{\rm LFO} = \sum_{i=L}^{N - M} \log p(y_{i+1:M} \,|\, y_{1:i}).
\end{equation}

The quantities $p(y_{i+1:M} \,|\, y_{1:i})$ can be computed with the help of the
posterior distribution $p(\theta \,|\, y_{1:i})$ of the parameters $\theta$
conditional on only the first $i-1$ observations of the time series:

\begin{equation}
\label{Lpred}
p(y_{i+1:M} \,| \, y_{1:i}) = 
  \int p(y_{i+1:M} \,| \, y_{1:i}, \theta) \, 
    p(\theta\,|\,y_{1:i}) \, {\rm d} \theta. 
\end{equation}

Most time series models have a non-factorizable likelihood, that is, $y_{i+1:M}$
depends on $y_{1:i}$ even after conditioning on $\theta$. As such, we cannot
simplify the integrand in (\ref{Lpred}) as would have been possible for
factorizable likelihoods. Instead, we always need to take $y_{1:i}$ into account
when computing the predictive density of $y_{i+1:M}$ (see
@buerkner:non-factorizable for more discussion on computing predictive densities
of non-factorizable models).

In practice, we will not be able to directly solve the integral in 
(\@ref(Lpred)), but instead have to use Monte-Carlo methods to approximate it.
Having obtained $S$ random draws $(\theta_{1:i}^{(1)}, \ldots, \theta_{1:i}^{(S)})$ 
from the posterior distribution $p(\theta\,|\,y_{1:i})$, we can estimate 
$p(y_{i+1:M} | y_{1:i})$ as

\begin{equation}
p(y_{i+1:M} \,|\, y_{1:i}) \approx \frac{1}{S}
\sum_{s=1}^S p(y_{i+1:M} \,|\, y_{1:i}, \theta_{1:i}^{(s)}).
\end{equation}

In this paper we use ELPD as a measure of predictive accuracy, but $M$-SAP (and
the approximations we introduce below) may also be based on other global
measures of accuracy such as the root mean squared error (RMSE) or the median
absolute deviation (MAD). The reason for our focus on ELPD is that it evaluates
a distribution rather than a point estimate (like the mean or median) to provide
a measure of out-of-sample predictive performance, which we see as favorable
from a Bayesian perspective [@vehtari2012]. The code we provide on GitHub
(https://github.com/paul-buerkner/LFO-CV-paper) is modularized to support
arbitrary measures of accuracy as long as they can be represented in a pointwise
manner, that is, as increments per observation. In Appendix C we also provide
additional simulation results using RMSE instead of ELPD.

## Approximate $M$-step-ahead predictions {#approximate-MSAP}

The equations above make use of posterior distributions from many
different fits of the model to different subsets of the data. To obtain
the predictive density $p(y_{i+1:M} \,|\, y_{1:i})$, a model is fit to
only the first $i$ data points, and we need to do this for every value of
$i$ under consideration (all $i \in \{L, \ldots, N - M\}$).
Below, we present a new algorithm to reduce the number of models that need
to be fit for the purpose of obtaining each of the densities 
$p(y_{i+1:M} \,|\, y_{1:i})$. This algorithm relies in a central manner on
Pareto smoothed importance sampling [@vehtari2017loo; @vehtari2019psis], which
we will briefly review next.

### Pareto smoothed importance sampling {#psis}

Importance sampling is a technique for computing expectations with
respect to some target distribution using an approximating proposal distribution
that is easier to draw samples from than the actual target. If $f(\theta)$ is
the target and $g(\theta)$ is the proposal distribution, we can write any
expectation of some function $h(\theta)$ with respect to $f$ as

\begin{equation}
\mathbb{E}_f[h(\theta)] = \int h(\theta) f(\theta) \,d\, \theta 
 = \frac{\int [h(\theta) f(\theta) / g(\theta)] g(\theta) \,d\, \theta}
    {\int [f(\theta) / g(\theta)] g(\theta) \,d\, \theta} 
 = \frac{\int h(\theta) r(\theta) g(\theta) \,d\, \theta}
    {\int r(\theta) g(\theta) \,d\, \theta}
\end{equation}

with importance ratios 
\begin{equation}
r(\theta) = \frac{f(\theta)}{g(\theta)}.
\end{equation}

Accordingly, if $\theta^{(s)}$ are $S$ random draws from $g(\theta)$, we can
approximate

\begin{equation}
\mathbb{E}_f[h(\theta)] \approx 
\frac{\sum_{s=1}^S h(\theta^{(s)}) r(\theta^{(s)})}{\sum_{s=1}^S r(\theta^{(s)})},
\end{equation}

provided that we can compute the raw importance ratios $r(\theta^{(s)})$ up to
some multiplicative constant. The raw importance ratios serve as weights on the
corresponding random draws in the approximation of the quantity of interest.

The main problem with this approach is that the raw importance ratios tend to
have large or infinite variance and results can be highly unstable. In order to
stabilize the computations, we can perform the additional step of regularizing
the largest raw importance ratios using the corresponding quantiles of the
generalized Pareto distribution fitted to the largest raw importance ratios.
This procedure is called Pareto smoothed importance sampling 
[PSIS; @vehtari2017loo; @vehtari2019psis, @loo2019] and has been demonstrated to
have a lower error and faster convergence rate than other commonly used
regularization techniques [@vehtari2019psis].

In addition, PSIS comes with a useful diagnostic to evaluate
the quality of the importance sampling approximation. The shape parameter $k$
of the generalized Pareto distribution fit to the largest importance ratios provides
information about the number of existing moments of the distribution of the
weights (smoothed ratios) and the actual importance sampling estimate. 
When $k<0.5$, the weight distribution has finite variance and the central limit
theorem ensures fast convergence of the importance sampling estimate with
increasing number of draws. This implies that approximate LOO-CV via PSIS is
highly accurate for $k<0.5$ [@vehtari2019psis]. For $0.5 \leq k < 1$, a
generalized central limit theorem holds, but the convergence rate drops quickly
as $k$ increases [@vehtari2019psis].  In practice, PSIS has been shown to be
relatively robust for $k < 0.7$ [@vehtari2017loo; @vehtari2019psis]. As such,
the default threshold is set to $0.7$ when performing 
PSIS LOO-CV [@vehtari2017loo;@loo2019].


### PSIS applied to $M$-step-ahead predictions {#psis-MSAP}

We now turn back to the task of estimating $M$-step-ahead performance for
time series models. First, we refit the model using the first $L$ observations
of the time series and then perform a single exact $M$-step-ahead prediction 
step for $p(y_{L+1:M} \,|\, y_{1:L})$. 
Recall that $L$ is the minimum number of observations we have deemed
acceptable for making predictions (setting $L=0$ means the first data point will
be predicted only based on the prior). Next, starting with $i = L + 1$, we
approximate each $p(y_{i+1:M} \,|\, y_{1:i})$ via

\begin{equation}
 p(y_{i+1:M} \,|\, y_{1:i}) \approx
   \frac{ \sum_{s=1}^S w_i^{(s)}\, p(y_{i+1:M} \,|\, \theta^{(s)})}
        { \sum_{s=1}^S w_i^{(s)}},
\end{equation}

where $w_i^{(s)}$ are the PSIS weights computed on the basis of the
raw importance ratios $r_i(\theta^{(s)})$ as explained in @vehtari2019psis, and $\theta^{(s)}$ are draws from the
posterior distribution based on the initial $L$ observations. 
To obtain $w_i^{(s)}$, we first compute the raw importance ratios

\begin{equation}
r_i^{(s)} = r_i(\theta^{(s)}) = \frac{f_i(\theta^{(s)})}{g(\theta^{(s)})} 
\propto \frac{
\prod_{j \in J} p(y_j \,|\, \,\theta^{(s)}) p(\theta^{(s)})
}{
\prod_{j \in J \backslash J_i} p(y_j \,|\, \,\theta^{(s)}) p(\theta^{(s)})
}
= \prod_{j \in J_i} p(y_j \,|\, \,\theta^{(s)}),
\end{equation}

with $J = \{1, \ldots, N\}$, and then stabilize them using PSIS as described
above. The index set $J_i$ contains the indices of all observations which are
part of the data for the model whose predictive performance we are trying to
approximate but not for the actually fitted model. Until a refit becomes
necessary (see below), we will have $J_i = \{L + 1, \ldots, i\}$.
For the starting value $i = L + 1$, $J_i$ is just $\{i\}$.

Continuing with the next observation, we gradually increase $i$ by $1$ 
(we move forward in time) and repeat the process. At some observation $i$, 
the variability of the importance ratios $r_i^{(s)}$ will become too large and
importance sampling will fail. We will refer to this particular value of $i$ as
$i^\star_1$. To identify the value of $i^\star_1$, we check for which value of
$i$ does the estimated shape parameter $k$ of the generalized Pareto
distribution first cross a certain threshold $\tau$ [@vehtari2019psis]. Only
then do we refit the model using the observations up to $i^\star_1$ and 
restart the process from there. Until the next refit we have 
$J_i = \{i^\star_1+1, \ldots, i \}$ for $i^\star_1 < i$, 
as the refitted model only contains the observations up to and including index
$i^\star_1$. An illustration of this procedure is shown in Figure
\@ref(fig:vis-msap). 

This bears some resemblance to Sequential Monte Carlo, also known as
particle or Monte Carlo filtering [e.g., @gordon1993; @kitagawa1996;
@doucet2000; @andrieu2010], in that applying SMC to state space models also
entails moving forward in time and using importance sampling to approximate the
next state from the information in the previous states [@kitagawa1996;
@andrieu2010]. However, in our case we are assuming we can sample from the
posterior distribution and instead concerned with estimating the model's
predictive performace. Unlike SMC, PSIS-LFO-CV also entails a full recomputation
of the model via Markov chain Monte Carlo (MCMC) once importance sampling fails.

In some cases we may only need to refit once and in other cases we will find a
value $i^\star_2$ that requires a second refitting, maybe an $i^\star_3$ that
requires a third refitting, and so on. We refit as many times as is required
(only when $k > \tau$) until we arrive at observation $i = N - M$. A detailed
description of the algorithm in the form of pseudo code is provided in Appendix
A. If the data contains multiple independent time series, the algorithm should
be applied to each of the time series separately, and the resulting ELPD values
can be summed up afterwards.

```{r vis-msap, fig.width=8, fig.height=3, fig.cap="Visualisation of PSIS approximated one-step-ahead predictions. Predicted observations are indicated by **X**. In the shown example, the model was last refit at the $i^\\star = 4$th observation."}
status_levels <- c("included", "included (PSIS)", "left out")
df <- data.frame(
  obs = rep(1:9, 3),
  i = factor(rep(3:5, each = 9)),
  Status = c(
    rep("included", 4), rep("left out", 5),
    rep("included", 4), rep("included (PSIS)", 1), rep("left out", 4),
    rep("included", 4), rep("included (PSIS)", 2), rep("left out", 3)
  )
) %>%
  mutate(Status = factor(Status, levels = status_levels))

msap_colors <- c(
  bayesplot::color_scheme_get("viridis")$light,
  bayesplot::color_scheme_get("viridis")$light_highlight,
  bayesplot::color_scheme_get("viridis")$dark
)

ggplot(df, aes(obs, i, fill = Status)) +
  geom_tile(height = 0.9, width = 1, col = "black") +
  annotate(
    'text', x = 5:7, y = c(1, 2, 3), 
    label = "X", parse = TRUE, 
    size = 10, color = "white"
  ) +
  labs(x = "Observation", y = "Predicted observation") +
  scale_x_continuous(breaks = 1:9) +
  scale_fill_manual(values = msap_colors) +
  bayesplot::theme_default() +
  NULL
```

Instead of moving forward in time, it is also possible to do 
PSIS-LFO-CV moving backwards in time. However, in that case the target posterior
is approximated by a distribution based on more observations and, as such, 
the proposal distribution is narrower than the target. This can result in 
highly influential importance weights more often than when 
the proposal is wider than the target, which is the case
for the forward PSIS-LFO-CV described above. In Appendix B we show 
that moving backwards indeed requires more refits than moving forward, and 
without any increase in accuracy. When we refer to the PSIS-LFO-CV algorithm 
in the main text we are referring to the forward version.

The threshold $\tau$ is crucial to the accuracy and speed of the PSIS-LFO-CV
algorithm. If $\tau$ is too large then we need fewer refits but accuracy will
suffer. If $\tau$ is too small then accuracy will be higher but many refits will
be required and the computation time may be impractical. Limiting the number of
refits without sacrificing too much accuracy is essential since almost all of
the computation time for exact cross-validation of Bayesian models is spent
fitting the models (not calculating the predictions). Therefore, any reduction
we can achieve in the number of refits essentially implies a proportional
reduction in the overall time required for cross-validation of Bayesian models.
We will come back to the issue of setting appropriate thresholds in Section
\@ref(simulations).

# Simulations {#simulations}

To evaluate the quality of the PSIS-LFO-CV approximation, we performed a
simulation study in which the following conditions were systematically varied: 

* The number $M$ of future observations to be predicted took on values of $M =
1$ and $M = 4$.
* The threshold $\tau$ of the Pareto $k$ estimates was varied between $k = 0.5$
to $k = 0.7$ in steps of $0.1$.
* Six different data generating models were evaluated, with linear and/or
quadratic terms and/or autoregressive terms of order 2 (see Figure 
\@ref(fig:simmodels) for an illustration).

In all cases the time series consisted of $N = 200$ observations and the minimal
number of observations required before make predictions was set to $L = 25$.
We ran $100$ simulation trials per condition.

```{r, include = FALSE}
seed <- 1234
set.seed(seed)
N <- 200
time <- seq_len(N)
stime <- scale_unit_interval(time)
models <- c(
  "constant", "linear", "quadratic",
  "AR2-only", "AR2-linear", "AR2-quadratic"
)

fits <- preds <- setNames(vector("list", length(models)), models)
for (m in names(fits)) {
   file <- paste0("models/fit_", m)
   fits[[m]] <- fit_model(model = m, N = N, seed = seed, file = file)
   pred <- posterior_predict(fits[[m]])
   preds[[m]] <- fits[[m]]$data %>% 
     mutate(       
       time = time, 
       stime = stime,
       Estimate = colMeans(pred), 
       Q5 = apply(pred, 2, quantile, probs = 0.05),
       Q95 = apply(pred, 2, quantile, probs = 0.95),
       model = m
    )
}
preds <- as_tibble(bind_rows(preds)) %>%
   mutate(model = factor(model, levels = models)) %>%
   select(-`I(stime^2)`)
```

```{r simmodels, fig.height=4, fig.cap="Illustration of the models used in the simulations. More details are provided in the text."}
ggplot(preds, aes(x = time, y = Estimate)) +
  facet_wrap(~model) +
  geom_smooth(aes(ymin = Q5, ymax = Q95), stat = "identity", size = 0.5) +
  geom_point(aes(y = y), size = 0.5) +
  labs(y = "y")
```

Autoregressive (AR) models are some of the most commonly used time series models. 
An AR(p) model -- an autoregressive model of order $p$ -- can be defined as

\begin{equation}
y_i = \eta_i + \varepsilon_i \quad \text{with} \quad
\varepsilon_i = \sum_{k = 1}^p \varphi_k \varepsilon_{i - k} + e_i,
\end{equation}
where $\eta_i$ is the linear predictor for the $i$th observation, $\varphi_k$
are the autoregressive parameters on the residuals $\varepsilon_i$, and $e_i$ are
pairwise independent errors, which are usually assumed to be normally
distributed with equal variance $\sigma^2$. The model implies a recursive
formula that allows for computing the right-hand side of the equation for
observation $i$ based on the values of the equations computed for previous
observations. Observations from an AR process are therefore not conditionally
independent by definition, but the likelihood is still factorizable because we
can write down a separate contribution for each observation [see
@buerkner:non-factorizable for more discussion on factorizability of statistical
models].

In the quadratic model with AR(2) effects (the most complex model in our
simulations), the true data generating process was defined as
\begin{equation}
y_i = b_0 + b_1 t + b_2 t^2 + \varepsilon_i \quad \text{with} \quad
\varepsilon_i = \varphi_1 \varepsilon_{i - 1} + \varphi_2 \varepsilon_{i - 2} + e_i,
\end{equation}
where $t$ is the time variable scaled to the unit interval, that is, $t = 0$ for
the smallest time point ($1$ in our simulations) and $t = 1$ for the largest
time point ($200$ in our simulations). Specifially, we set the true regression
coefficients to the values of $b_0 = 0$, $b_1 = 17$, $b_2 = 25$, and the true
autocorrelation parameters to $\varphi_1 = 0.5$, and $\varphi_2 = 0.3$ (see
Figure \@ref(fig:simmodels) for an illustration). The choices of the regression
coefficients were done so that neither the linear nor quadratic term dominates
the other within the specified time frame. The values of the autocorrelation
parameters were set to represent typical positively autocorrelated data.
In the simulation conditions without linear and/or quadratic and/or 
AR(2) terms, the corresponding true parameters were simply fixed to zero.

In addition to exact and approximate LFO-CV, we also computed approximate LOO-CV
for comparison. This is not because we think LOO-CV is a generally appropriate
approach for time series models, but because, in the absence of any approximate
LFO-CV method, researchers may have used approximate LOO-CV for time series
models in the past simply because it was the only available option. 
Demonstrating that LOO-CV is a biased estimate of LFO-CV underscores 
the importance of developing methods better suited for the task.

All simulations were done in R [@R2018] using the brms package [@brms1;
@brms2] together with the probabilistic programming language Stan
[@carpenter2017; @rstan2019] for model fitting, the loo R package
[@loo2019] for the PSIS computations, and several tidyverse R packages
[@tidyverse] for data processing. The full code and all results are available on
Github (https://github.com/paul-buerkner/LFO-CV-paper).

## Results {#sim_results}

```{r}
# extractor helper functions
get_elpd <- function(x) {
  summarize_elpds(x)[1]
}
get_refits <- function(x) {
  sum(attr(x, "refits"), na.rm = TRUE)
}

# pretty factor labels
mlevels <- c(
  "constant", "linear", "quadratic",
  "AR2-only", "AR2-linear", "AR2-quadratic"
)
tau_levels <- TeX(paste0("$\\tau$ = ", c(0.5, 0.6, 0.7)))

elpd_colors <- c(
  loo = bayesplot::color_scheme_get("viridis")$mid,
  lfo_bw = bayesplot::color_scheme_get("viridis")$light,
  lfo_fw = bayesplot::color_scheme_get("viridis")$dark,
  lfo_cb = bayesplot::color_scheme_get("viridis")$mid
)

# data preparation
lfo_sims <- read_rds("results/lfo_sims.rds") %>%
  as_tibble() %>%
  filter(lengths(res) > 0) %>%
  mutate(
    model = factor(model, levels = mlevels),
    tau = factor(k_thres, labels = tau_levels),
    # compute sum of pointwise ELPD values
    elpd_loo = map_dbl(res, ~ .$loo_cv$estimates["elpd_loo", 1]),
    elpd_exact_lfo = map_dbl(res, ~ get_elpd(.$lfo_exact_elpds)),
    elpd_approx_fw_lfo = map_dbl(res, ~ get_elpd(.$lfo_approx_fw_elpds)),
    elpd_approx_bw_lfo = map_dbl(res, ~ get_elpd(.$lfo_approx_bw_elpds)),
    # compute differences to the exact values
    elpd_diff_bw_lfo = elpd_approx_bw_lfo - elpd_exact_lfo,
    elpd_diff_fw_lfo = elpd_approx_fw_lfo - elpd_exact_lfo,
    elpd_diff_loo = elpd_loo - elpd_exact_lfo,
    nrefits_bw = map_dbl(res, ~ get_refits(.$lfo_approx_bw_elpds)),
    nrefits_fw = map_dbl(res, ~ get_refits(.$lfo_approx_fw_elpds)),
    rel_nrefits_bw = nrefits_bw / (N - L),
    rel_nrefits_fw = nrefits_fw / (N - L)
    # the combined mode is not yet fully implemented
    # elpd_approx_cb_lfo  = map_dbl(res, ~ get_elpd(.$lfo_approx_cb_elpds)),
    # elpd_diff_cb_lfo = elpd_approx_cb_lfo - elpd_exact_lfo,
    # nrefits_cb = map_dbl(res, ~ get_refits(.$lfo_approx_cb_elpds)),
    # rel_nrefits_cb = nrefits_cb / (N - L)
  ) %>%
  select(-res)
```

In this section we focus on the ELPD as a measure out-of-sample predictive
performance for reasons outlined in Section \ref{m-sap}. In Appendix C, 
we provide additional simulation results for the RMSE. 

Results of the 1-SAP simulations are visualized in Figure \@ref(fig:1sap).
Comparing the columns of Figure \@ref(fig:1sap), it is clearly visible that the
the accuracy of the PSIS approximation is independent of the threshold $\tau$
when $\tau$ is within the interval $[0.5,0.7]$ motivated in \@ref(psis) [this
would not be the case if $\tau$ was allowed to be larger; @vehtari2019psis]. For
all conditions, the PSIS-LFO-CV approximation is highly accurate, that is, both
unbiased and low in variance around the corresponding exact LFO-CV value
(represented by the dashed line in Figure \@ref(fig:1sap)). The proportion of
observations at which refitting the model was required did not exceed $3\%$
under any of the conditions and only increased minimally when decreasing $\tau$
(see Table \@ref(tab:refits)). At least for the models investigated in our
simulations, using $\tau = 0.7$ seems to be sufficient for achieving high
accuracy and as such there is no need to lower the threshold below that value.
As expected, LOO-CV (the lighter histograms in Figure \@ref(fig:1sap)) is a
biased estimate of the 1-SAP performance for all non-constant models, in
particular for models with a trend in the time series. More precisely, LOO-CV is
positively biased, which implies that it systematically overestimates $M$-SAP
performance of time series models.

```{r 1sap, fig.height=8, fig.cap="Simulation results of 1-step-ahead predictions. Histograms are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. The black dashed lines indicates the exact LFO-CV result."}
lfo_sims %>% 
  filter(is.na(B), M == 1) %>% 
  select(elpd_diff_fw_lfo, elpd_diff_loo, model, tau) %>%
  gather("Mode", "elpd_diff", starts_with("elpd_diff")) %>%
  ggplot(aes(x = elpd_diff, y = ..density.., fill = Mode)) +
  facet_grid(
    model ~ tau, scales = "free_y",
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7) +
  scale_fill_manual(
    values = unname(elpd_colors[c("lfo_fw", "loo")]),
    labels = c("PSIS-LFO-CV", "PSIS-LOO-CV"),
  ) +
  labs(x = "ELPD difference of approximate and exact LFO-CV", y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    legend.position = "bottom"
  ) +
  NULL
```

```{r refits, cache=FALSE}
lfo_sims %>% 
  filter(is.na(B)) %>%
  select(rel_nrefits_fw, model, M, k_thres) %>%
  group_by(model, M, k_thres) %>%
  summarise(rel_nrefits = round(mean(rel_nrefits_fw), 2)) %>%
  ungroup() %>%
  spread("model", "rel_nrefits") %>%
  mutate(M = ifelse(duplicated(M), "", M)) %>%
  rename(`$\\tau$` = "k_thres") %>%
  kable(
    caption = "Mean proportions of required refits for PSIS-LFO-CV.",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  footnote(
    general = "Note: Results are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. Abbreviations: $\\\\tau$ = threshold of the Pareto $k$ estimates; $M$ = number of predicted future observations.",
    general_title = "",
    threeparttable = TRUE,
    escape = FALSE
  )
```

Results of the 4-SAP simulations are visualized in Figure \@ref(fig:4sap).
Comparing the columns of Figure \@ref(fig:4sap), it is again clearly visible
that the accuracy of the PSIS approximation is independent of the threshold
$\tau$. The proportion of observations at which refitting the model was required
did not exceed $3\%$ under any condition and only increased minimally when
decreasing $\tau$ (see Table \@ref(tab:refits)). In light of the corresponding
1-SAP results presented above, this is not surprising because the procedure for
determining the necessity of a refit is independent of $M$ (see Section
\@ref(approximate-MSAP)). PSIS-LOO-CV is not displayed in Figure \@ref(fig:4sap) as
the number of observations predicted at each step (4 vs. 1) makes 4-SAP LFO-CV
and LOO-CV incomparable.

When looking at the bottom three rows of Figure \@ref(fig:4sap), we see that
there is a lot of variation around the true 4-SAP ELPD value across simulation
trials, although in expectation PSIS-LFO-CV is still unbiased. This variation is
caused by the fact that we predict _four_ observations into the future using AR
models of at most order _two_. Predictions for the 3rd and 4th observations into
the future will therefore have much greater uncertainty and so will the
PSIS-LFO-CV approximation. This is not a limitation of the method, but rather an
artifact of the limited predictive capabilities of these particular models.

```{r 4sap, fig.height=8, fig.cap="Simulation results of 4-step-ahead predictions. Histograms are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. The black dashed lines indicates the exact LFO-CV result."}
lfo_sims %>% 
  filter(is.na(B), M == 4) %>%
  select(elpd_diff_fw_lfo, model, tau) %>%
  ggplot(aes(x = elpd_diff_fw_lfo, y = ..density..)) +
  facet_grid(
    model ~ tau, scales = "free_y",
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7, fill = unname(elpd_colors["lfo_fw"])) +
  labs(x = "ELPD difference of approximate and exact LFO-CV", y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  xlim(c(-30, 30)) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    legend.position = "bottom"
  ) +
  NULL
```

# Case Studies {#case-studies}

## Annual measurements of the level of Lake Huron {#case-LH}

To illustrate the application of PSIS-LFO-CV for estimating expected $M$-SAP
performance, we will fit a model for 98 annual measurements of the water level
(in feet) of [Lake Huron](https://en.wikipedia.org/wiki/Lake_Huron) from the
years 1875--1972. This data set is found in the *datasets* R package, which is
installed automatically with R [@R2018]. The time series shows rather strong
autocorrelation and some downward trend towards lower water levels for later points
in time. Figure \@ref(fig:lake-huron) shows the observed time series of water 
levels as well as predictions from a fitted AR(4) model.

```{r}
data("LakeHuron")
N <- length(LakeHuron)
df <- data.frame(
  y = as.numeric(LakeHuron),
  year = as.numeric(time(LakeHuron)),
  time = 1:N
) 
```

```{r fit_lh, results = "hide"}
fit_lh <- brm(
  y | mi() ~ 1, 
  data = df, 
  autocor = cor_ar(~time, p = 4), 
  prior = prior(normal(0, 0.5), class = "ar"),
  chains = 2, warmup = 1000, iter = 5000,
  control = list(adapt_delta = 0.99),
  seed = 5838296, file = "models/fit_lh"
)
```

```{r lake-huron, fig.cap="Water Level in Lake Huron (1875-1972). Black points are observed data. The blue line represents mean predictions of an AR(4) model with 90% prediction intervals shown in gray.", fig.height=3}
preds <- posterior_predict(fit_lh)
preds <- cbind(
  Estimate = colMeans(preds), 
  Q5 = apply(preds, 2, quantile, probs = 0.05),
  Q95 = apply(preds, 2, quantile, probs = 0.95)
)

ggplot(cbind(df, preds), aes(x = year, y = Estimate)) +
  geom_smooth(aes(ymin = Q5, ymax = Q95), stat = "identity", size = 0.5) +
  geom_point(aes(y = y)) + 
  labs(y = "Water Level (ft)", x = "Year")
```

```{r}
L <- 20
k_thres <- 0.7
loo_lh <- loo(log_lik(fit_lh)[, (L + 1):N])
```

Based on this data and model, we will illustrate the use of PSIS-LFO-CV to
provide estimates of $1$-SAP and $4$-SAP when leaving out all future values. To
allow for reasonable predictions, we will require at least $L = 20$ historical
observations (20 years) to make predictions. Further, we set a threshold of
$\tau =$ `r k_thres` for the Pareto $k$ estimates that indicate when refitting
becomes necessary. Our fully reproducible analysis of this case study can be
found on GitHub (https://github.com/paul-buerkner/LFO-CV-paper).

```{r}
lh_elpd_1sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = 1, L = L, 
  file = "results/lh_elpd_1sap_exact.rds"
)
lh_elpd_1sap_approx <- compute_lfo(
  fit_lh, type = "approx", M = 1, L = L,
  k_thres = k_thres, mode = "forward",
  file = "results/lh_elpd_1sap_approx_fw.rds"
)
lh_nrefits <- get_refits(lh_elpd_1sap_approx)
```

```{r}
lh_elpd_4sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = 4, L = L, 
  file = "results/lh_elpd_4sap_exact.rds"
)
lh_elpd_4sap_approx <- compute_lfo(
  fit_lh, type = "approx", M = 4, L = L, 
  k_thres = k_thres, mode = "forward",
  file = "results/lh_elpd_4sap_approx_fw.rds"
)
```

We start by computing exact and PSIS-approximated LFO-CV of 1-SAP. The computed
ELPD values are
${\rm ELPD}_{\rm exact} =$ `r fmt(get_elpd(lh_elpd_1sap_exact), 2)` and 
${\rm ELPD}_{\rm approx} =$ `r fmt(get_elpd(lh_elpd_1sap_approx), 2)`, which are
almost identical. Not only is the overall ELPD estimated accurately but so are
all of the pointwise ELPD contributions (see the left panel of Figure
\@ref(fig:lh-pw-elpd)). In comparison, PSIS-LOO-CV returns
${\rm ELPD}_{\rm loo} =$ `r fmt(loo_lh$estimates[1, 1], 1)`,
overestimating the predictive performance and as suggested by our simulation
results for stationary autoregessive models (see fourth row of Figure
\@ref(fig:1sap)). Plotting the Pareto $k$ estimates reveals that the model had
to be refit `r lh_nrefits` times, out of a total of $N - L =$ `r N - L` predicted
observations (see Figure \@ref(fig:lh-pareto-k)). On average, this means one
refit every `r fmt((N - L) / lh_nrefits, 1)` observations, which implies a drastic
speed increase compared to exact LFO-CV.

Performing LFO-CV for 4-SAP, we obtained ${\rm ELPD}_{\rm exact} =$ 
`r fmt(get_elpd(lh_elpd_4sap_exact), 2)` and ${\rm ELPD}_{\rm approx} =$ 
`r fmt(get_elpd(lh_elpd_4sap_approx), 2)`, which are again very similar.
In general, as $M$ increases, the approximation will tend to become more variable
around the true value in absolute ELPD units because the ELPD increment of each 
observation will be based on more and more observations (see also Section 
\@ref(simulations)). For this example, we see some considerable differences in
the pointwise ELPD contributions of specific observations which were hard to
predict accurately by the model (see the right panel of Figure
\@ref(fig:lh-pw-elpd)). This is to be expected because predicting $M$ steps ahead
using an AR model will yield highly uncertain predictions if most of the
autocorrelation happens at lags smaller than $M$ (see also the bottom rows in
Figure \ref{fig:4sap}). For such a model, it may be ill-advised to evaluate
predictions too far into the future, at least when using the approximate methods
presented in this paper. Since, for a constant threshold $\tau$, the importance
weights are the same independent of $M$, the Pareto $k$ estimates are the same
for $4$-SAP and $1$-SAP.

```{r lh-pw-elpd, warning=FALSE, fig.height=3, fig.cap="Pointwise exact vs. PSIS-approximated ELPD contributions for 1-SAP (left) and 4-SAP (right) for the Lake Huron model. A threshold of $\\tau = 0.7$ was used for the Pareto $k$ estimates. $M$ is the number of predicted future observations."}
lh_pw_elpd <- tibble(
  elpd_exact = na.omit(lh_elpd_1sap_exact),
  elpd_approx = na.omit(lh_elpd_1sap_approx),
  k = na.omit(attributes(lh_elpd_1sap_approx)$ks),
  M = "M = 1"
) %>% bind_rows(
  tibble(
    elpd_exact = na.omit(lh_elpd_4sap_exact),
    elpd_approx = na.omit(lh_elpd_4sap_approx),
    k = na.omit(attributes(lh_elpd_4sap_approx)$ks),
    M = "M = 4"
  )
)
ggplot(lh_pw_elpd, aes(elpd_exact, elpd_approx)) +
  facet_wrap(facets = "M", nrow = 1, ncol = 2, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate ELPD", x = "Exact ELPD") + 
  theme_bw()
```

```{r lh-pareto-k, fig.height=2.5, fig.cap="Pareto $k$ estimates for PSIS-LFO-CV of the Lake Huron model. The dotted red line indicates the threshold at which the refitting was necessary."}
ks <- na.omit(attributes(lh_elpd_1sap_approx)$ks)
ids <- (L + 1):N
plot_ks(ks, ids, k_thres = k_thres)
```


## Annual date of the cherry blossoms in Japan {#case-CB}

```{r}
cherry <- read.csv("data/cherry_blossoms.csv")
cherry_temp <- cherry[!is.na(cherry$temp), ]
cherry_doy <- cherry[!is.na(cherry$doy), ]
```

The cherry blossom in Japan is a famous natural phenomenon occurring once every
year during spring. As the climate changes so does the annual date of the cherry
blossom [@aono2008; @aono2010]. The most complete reconstruction available to
date contains data between `r min(cherry$year)` AD and `r max(cherry$year)` AD
[@aono2008; @aono2010] and is available online
(http://atmenv.envi.osakafu-u.ac.jp/aono/kyophenotemp4/).

In this case study, we will predict the annual date of the cherry
blossom using an approximate Gaussian process model [@solin2014; @RiutortMayol2019] to
provide flexible non-linear smoothing of the time series. A visualisation of
both the data and the fitted model is provided in Figure
\@ref(fig:cherry-blossom). While the time series appears rather stable across
earlier centuries, with substantial variation across consecutive years, there
are some clearly visible trends in the data. Particularly in more recent years,
the cherry blossom has tended to happen much earlier than before, which may 
be a consequence of changes in the climate [@aono2008; @aono2010].

Based on this data and model, we will illustrate the use of PSIS-LFO-CV to
provide estimates of $1$-SAP and $4$-SAP leaving out all future values. To allow
for reasonable predictions of future values, we will require at least $L = 100$
historical observations (100 years) to make predictions. Further, we set a
threshold of $\tau =$ `r k_thres` for the Pareto $k$ estimates to determine when
refitting becomes necessary. Our fully reproducible analysis of this case study 
can be found on GitHub (https://github.com/paul-buerkner/LFO-CV-paper).

```{r fit_cb}
fit_cb <- brm(
  formula = bf(doy ~ gp(year, k = 20, c = 5/4)), 
  data = cherry_doy, 
  prior = prior(normal(0, 0.1), class = lscale, coef = gpyear),
  chain = 2, warmup = 4000, iter = 7000, inits = 0,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  seed = 5838234, file = "models/fit_cb" 
)
```

```{r cherry-blossom, fig.height=3, fig.cap="Day of the cherry blossom in Japan (812-2015). Black points are observed data. The blue line represents mean predictions of a thin-plate spline model with 90% regression intervals shown in gray."}
me_cb <- marginal_effects(fit_cb, probs = c(0.05, 0.95))
plot(me_cb, points = TRUE, plot = FALSE)[[1]] +
  labs(x = "Year", y = "Day of cherry blossom")
```

```{r}
N <- NROW(cherry_doy)
L <- 100
k_thres <- 0.7

cb_elpd_1sap_exact <- compute_lfo(
  fit_cb, type = "exact", M = 1, L = L, 
  factorizes = TRUE,
  file = "results/cb_elpd_1sap_exact.rds"
)
cb_elpd_1sap_approx <- compute_lfo(
  fit_cb, type = "approx", M = 1, L = L, 
  k_thres = k_thres, mode = "forward", factorizes = TRUE,
  file = "results/cb_elpd_1sap_approx_fw.rds"
)
cb_nrefits <- get_refits(cb_elpd_1sap_approx)

loo_cb <- loo(fit_cb, newdata = cherry_doy[-seq_len(L), ])
```

```{r}
cb_elpd_4sap_exact <- compute_lfo(
  fit_cb, type = "exact", M = 4, L = L, 
  file = "results/cb_elpd_4sap_exact.rds"
)
cb_elpd_4sap_approx <- compute_lfo(
  fit_cb, type = "approx", M = 4, L = L,
  k_thres = k_thres, mode = "forward", factorizes = TRUE,
  file = "results/cb_elpd_4sap_approx_fw.rds"
)
```

We start by computing exact and PSIS-approximated LFO-CV of 1-SAP. We compute
${\rm ELPD}_{\rm exact} =$ `r fmt(get_elpd(cb_elpd_1sap_exact), 1)` and 
${\rm ELPD}_{\rm approx} =$ `r fmt(get_elpd(cb_elpd_1sap_approx), 1)`, 
which are highly similar. As shown in the left panel of Figure
\@ref(fig:cb-pw-elpd), the pointwise ELPD contributions are highly accurate,
with no outliers. The approximation has worked well for all observations. 
PSIS-LFO-CV performs much better than PSIS-LOO-CV
(${\rm ELPD}_{\rm approx} =$  `r fmt(loo_cb$estimates[1, 1], 1)`), which
overestimates the predictive performance. Plotting the Pareto $k$ estimates
reveals that the model had to be refit `r cb_nrefits` times, out of a total of $N -
L =$ `r N - L` predicted observations (see Figure \@ref(fig:cb-pareto-k)). On
average, this means one refit every `r fmt((N - L) / cb_nrefits, 1)` observations,
which implies a drastic speed increase as compared to exact LFO-CV.

Performing LFO-CV of 4-SAP, we compute ${\rm ELPD}_{\rm exact} =$ 
`r fmt(get_elpd(cb_elpd_4sap_exact), 1)` and ${\rm ELPD}_{\rm approx} =$
`r fmt(get_elpd(cb_elpd_4sap_approx), 1)`, which are again similar but not as close
as the corresponding 1-SAP results. This is to be expected as the uncertainty of
PSIS-LFO-CV increases for increasing $M$ (see Section \@ref(simulations)). As
displayed in the right panel of Figure \@ref(fig:cb-pw-elpd), the pointwise ELPD
contributions are highly accurate in most cases, with a few small outliers in
both directions. For constant threshold $\tau$, the importance weights are the
same independent of $M$, so the Pareto $k$ estimates are the same for $4$-SAP
and $1$-SAP.

```{r cb-pw-elpd, warning=FALSE, fig.height=3, fig.cap="Pointwise exact vs. PSIS-approximated ELPD contributions of 1-SAP (left) and 4-SAP (right) for the cherry blossom model. A threshold of $\\tau = 0.7$ was used for the Pareto $k$ estimates. $M$ is the number of predicted future observations."}
cb_pw_elpd <- tibble(
  elpd_exact = na.omit(cb_elpd_1sap_exact),
  elpd_approx_fw = na.omit(cb_elpd_1sap_approx),
  k = na.omit(attributes(cb_elpd_1sap_approx)$ks),
  M = "M = 1"
) %>% bind_rows(
  tibble(
    elpd_exact = na.omit(cb_elpd_4sap_exact),
    elpd_approx_fw = na.omit(cb_elpd_4sap_approx),
    k = na.omit(attributes(cb_elpd_4sap_approx)$ks),
    M = "M = 4"
  )
)
ggplot(cb_pw_elpd, aes(elpd_exact, elpd_approx_fw)) +
  facet_wrap(facets = "M", nrow = 1, ncol = 2, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate ELPD", x = "Exact ELPD") + 
  theme_bw()
```

```{r cb-pareto-k, fig.height=2.5, fig.cap="Pareto $k$ estimates for PSIS-LFO-CV of the cherry blossom model. The dotted red line indicates the threshold at which the refitting was necessary."}
ks <- na.omit(attributes(cb_elpd_1sap_approx)$ks)
ids <- (L + 1):N
plot_ks(ks, ids, k_thres = k_thres)
```


# Conclusion {#discussion}

We proposed, evaluated, and demonstrated PSIS-LFO-CV, a method for approximating
cross-validation of Bayesian time series models. PSIS-LFO-CV is intended to be
used when the prediction task is predicting future values based solely on past
values, in which case leave-one-out cross-validation is inappropriate.
Within the set of such prediction tasks, we can choose the number $M$ of future
observations to be predicted. For a set of common time series models, we
established via simulations that PSIS-LFO-CV is an unbiased approximation
of exact LFO-CV if we choose the threshold $\tau$ of the Pareto $k$ estimates to
not be larger than $\tau = 0.7$. That is, PSIS-LFO-CV does not require a smaller
(stricter) threshold than PSIS-LOO-CV to achieve satistfactory accuracy.

By nature of the approximated $M$-step-ahead predictions, the computation time
of PSIS-LFO-CV still increases linearily with the number of observations $N$.
However, in our numerical experiments, we were able to reduce to computation
time by a factor of roughly 25 to 100 compared to exact LFO-CV, which is
enough to make LFO-CV realistic for many applications.

Lastly, we want to briefly note that LFO-CV can also be used to compute marginal
likelihoods. Using basic rules of conditional probability, we can factor the
log marginal likelihood as

\begin{equation}
\log p(y) = \sum_{i=1}^N \log p(y_i \,|\, y_{1:(i-1)}).
\end{equation}

This is exactly the ELPD of 1-SAP if we set $L = 0$, that is if we
choose to predict *all* observations using their respective past (the very
first observation is only predicted from the prior). As such, marginal
likelihoods may be approximated using PSIS-LFO-CV. Although this approach is
unlikely to be more efficient than methods specialized for computing marginal
likelihoods (e.g., bridge sampling [@meng1996; @meng2002; @gronau2017]), it may
be a noteworthy option if for some reason other methods fail.

# Acknowledgments

We thank Daniel Simpson, Shira Mitchell, and M\r{a}ns Magnusson for helpful
comments and discussions on earlier versions of this paper. We also acknowledge
the computational resources provided by the Aalto Science-IT project.

\newpage

# Appendix {-}

## Appendix A: Pseudo code for PSIS LFO-CV {-}

The R flavored pseudo code below provides a description of the proposed
PSIS-LFO-CV algorithm when leaving out all future values. See https://github.com/paul-buerkner/LFO-CV-paper for the actual R code.

```{r, tidy=FALSE, eval=FALSE, echo = TRUE}
# Approximate Leave-Future-Out Cross-Validation (LFO-CV)
# Arguments:
#   model: the fitted time series model based on the complete data
#   data: the complete data set
#   M: number of steps to be predicted into the future
#   L: minimal number of observations necessary to make predictions
#   tau: threshold of the Pareto-k-values
# Returns:
#   PSIS approximated ELPD value of LFO-CV
PSIS_LFO_CV = function(model, data, M, L, tau) {
  N = number_of_rows(data)
  S = number_of_draws(model)
  out = vector(length = N)
  # refit the model using the first L observations
  i_star = L
  model_star = update(model, data = data[1:L, ])
  out[L] = exact_ELPD(model_star, data = data[(L + 1):(L + M), ])
  # loop over all observations at which to perform predictions
  for (i in (L + 1):(N - M)) {
    PSIS_object = PSIS(model_star, data = data[(i_star + 1):i , ])
    k = pareto_k_values(PSIS_object)
    if (k > tau) {
      # refitting the model is necessary
      i_star = i
      model_star = update(model_star, data = data[1:i, ])
      out[i] = exact_ELPD(model_star, data = data[(i + 1):(i + M), ])
    } else {
      # PSIS approximation is possible
      log_PSIS_weights = log_weights(PSIS_object)
      out[i] = approx_ELPD(model_star, data = data[(i + 1):(i + M), ],
                           log_weights = log_PSIS_weights)
    }
  }
  return(sum(out))
}
```


## Appendix B: Backward PSIS-LFO-CV {-}

Instead of moving forward in time, that is, starting our predictions from the
$L$th observation, we may also move backwards, a procedure to which we will
refer to as backward PSIS-LFO-CV. Starting with $i = N - M$, we approximate each
$p(y_{i+1:M} \,|\, y_{1:i})$ via

\begin{equation}
 p(y_{i+1:M} \,|\, y_{1:i}) \approx
   \frac{ \sum_{s=1}^S w_i^{(s)}\, p(y_{i+1:M} \,|\, \theta^{(s)})}
        { \sum_{s=1}^S w_i^{(s)}},
\end{equation}

where $w_i^{(s)}$ are the PSIS weights and $\theta^{(s)}$ are draws from the
posterior distribution based on _all_ observations. To obtain $w_i^{(s)}$, we
first compute the raw importance ratios

\begin{equation}
r_i^{(s)} = r_i(\theta^{(s)}) = \frac{f_i(\theta^{(s)})}{g(\theta^{(s)})} 
\propto \frac{
\prod_{j \in J \backslash J_i} p(y_j \,|\, \,\theta^{(s)}) p(\theta^{(s)})
}{
\prod_{j \in J} p(y_j \,|\, \,\theta^{(s)}) p(\theta^{(s)})
}
= \frac{1}{\prod_{j \in J_i} p(y_j \,|\, \,\theta^{(s)})},
\end{equation}

with $J = \{1, \ldots, N\}$, and then stabilize them using PSIS as described
above. The index set $J_i$ contains the indices of all observations which are
part of the data for the model being fitted but not for the model whose
predictive performance we are trying to approximate. That is, for the starting
value $i = N - M$, we have $J_i = \{i+1, \ldots, N\}$. This approach to
computing importance ratios is a generalization of the approach used in
PSIS-LOO-CV, where only a single observation is left out at a time and thus 
$J_i = \{i\}$ for all $i$.

Starting from $i = N - M$, we gradually *decrease* $i$ by $1$ (i.e., we move
backwards in time) and repeat the process. At some observation $i$, the
variability of the importance ratios $r_i^{(s)}$ will become too large and
importance sampling fails. We will refer to this particular value of $i$ as
$i^\star_1$. To identify the value of $i^\star_1$, we check for which value of
$i$ does the estimated shape parameter $k$ of the generalized Pareto
distribution first cross a certain threshold $\tau$ [@vehtari2019psis]. Only
then do we refit the model using only observations up to $i^\star_1$ and then
restart the process. Until the next refit, we thus have $J_i = \{i+1, \ldots,
i^\star_1 \}$ for $i < i^\star_1$, as the refitted model only contains the
observations up to index $i^\star_1$. An illustration of this procedure is shown
in Figure \@ref(fig:vis-msap-bw). In some cases we may only need to refit once
and in other cases we will find a value $i^\star_2$ that requires a second
refitting, maybe an $i^\star_3$ that requires a third refitting, and so on. We
repeat the refitting as many times as is required (only if $k > \tau$) until we
arrive at $i = L$. Recall that $L$ is the minimum number of observations we have
deemed acceptable for making predictions.

In forward PSIS-LFO-CV, we have seen a threshold of $\tau = 0.7$ to be
sufficient for achieving satisfactory accuracy. For backward PSIS-LFO-CV, $\tau$
likely has to be smaller. More precisely, we can expect an appropriate threshold
for the backward mode to be somewhere between $0.5 \leq \tau \leq 0.7$. It is
unlikely to be as high as the $\tau = 0.7$ default used for PSIS-LOO-CV because
there will be more dependence in the errors in the case of backward PSIS-LFO-CV.
If there is a large error when leaving out the $i$th observation, then there is
likely to also be a large error when leaving out observations $i, i-1, i-2,
\ldots$ until a refit is performed. This means that highly influential
observations (ones with a large $k$ estimate) are likely to have stronger
effects on the total estimate for backward LFO-CV than for LOO-CV.

```{r vis-msap-bw, fig.width=8, fig.height=3, fig.cap="Visualisation of approximate one-step-ahead predictions using backward PSIS-LFO-CV. Predicted observations are indicated by **X**. In the shown example, the model was last refit at the $i^\\star = 4$th observation."}
status_levels <- c("included", "left out", "left out (PSIS)")
df <- data.frame(
  obs = rep(1:9, 3),
  i = factor(rep(3:5, each = 9)),
  Status = c(
    rep("included", 2), rep("left out (PSIS)", 2), rep("left out", 5),
    rep("included", 3), rep("left out (PSIS)", 1), rep("left out", 5),
    rep("included", 4), rep("left out", 5)
  )
) %>%
  mutate(Status = factor(Status, levels = status_levels))

msap_colors <- c(
  bayesplot::color_scheme_get("viridis")$light,
  bayesplot::color_scheme_get("viridis")$dark,
  bayesplot::color_scheme_get("viridis")$mid_highlight
)

ggplot(df, aes(obs, i, fill = Status)) +
  geom_tile(height = 0.9, width = 1, col = "black") +
  annotate(
    'text', x = 3:5, y = c(1, 2, 3), 
    label = "X", parse = TRUE, 
    size = 10, color = "white"
  ) +
  labs(x = "Observation", y = "Predicted observation") +
  scale_x_continuous(breaks = 1:9) +
  scale_fill_manual(values = msap_colors) +
  bayesplot::theme_default() +
  NULL
```

The simulation results comparing backward to forward PSIS-LFO-CV can be found in
Figure \ref{fig:1sap-both} for $1$-SAP and in Figure \ref{fig:4sap-both} for
$4$-SAP. As visible in both figures, backward PSIS-LFO-CV requires a lower
$\tau$ threshold than forward PSIS-LFO-CV in order to be accurate ($\tau = 0.6$
vs. $\tau = 0.7$). Otherwise, it may have a small positive bias. Further, as can
be seen in Table \ref{tab:refits-both}, backward PSIS-LFO-CV requires
considerably more refits then forward PSIS-LFO-CV. Together, this indicates
that, in expectation, backward PSIS-LFO-CV is inferior to forward PSIS-LFO-CV.

```{r 1sap-both, fig.height=8, fig.cap="Simulation results of 1-step-ahead predictions for both forward and backward PSIS-LFO-CV. Histograms are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. The black dashed lines indicates the exact LFO-CV result."}
lfo_sims %>% 
  filter(is.na(B), M == 1) %>% 
  select(starts_with("elpd_diff"), model, tau) %>%
  gather("Mode", "elpd_diff", starts_with("elpd_diff")) %>%
  ggplot(aes(x = elpd_diff, y = ..density.., fill = Mode)) +
  facet_grid(
    model ~ tau, scales = "free_y",
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7) +
  scale_fill_manual(
    values = unname(elpd_colors[c("lfo_bw", "lfo_fw", "loo")]),
    labels = c(
      "PSIS-LFO-CV (backward)", 
      "PSIS-LFO-CV (forward)",
      "PSIS-LOO-CV"
    )
  ) +
  labs(x = "ELPD difference of approximate and exact LFO-CV", y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    legend.position = "bottom"
  ) +
  NULL
```

```{r 4sap-both, fig.height=8, fig.cap="Simulation results of 4-step-ahead predictions for both forward and backward PSIS-LFO-CV. Histograms are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. The black dashed lines indicates the exact LFO-CV result."}
lfo_sims %>% 
  filter(is.na(B), M == 4) %>%
  select(starts_with("elpd_diff"), model, tau, -elpd_diff_loo) %>%
  gather("Mode", "elpd_diff", starts_with("elpd_diff")) %>%
  ggplot(aes(x = elpd_diff, y = ..density.., fill = Mode)) +
  facet_grid(
    model ~ tau, scales = "free_y",
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7) +
  scale_fill_manual(
    values = unname(elpd_colors[c("lfo_bw", "lfo_fw")]),
    labels = c(
      "PSIS-LFO-CV (backward)", 
      "PSIS-LFO-CV (forward)"
    )
  ) +
  labs(x = "ELPD difference of approximate and exact LFO-CV", y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  xlim(c(-30, 30)) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    legend.position = "bottom"
  ) +
  NULL
```


```{r refits-both, cache=FALSE}
lfo_sims %>% 
  filter(is.na(B)) %>%
  select(starts_with("rel_nrefits"), model, M, k_thres) %>%
  gather("Mode", "rel_nrefits", starts_with("rel_nrefits")) %>%
  mutate(
    Mode = case_when(
      grepl("_fw$", Mode) ~ "forward",
      grepl("_bw$", Mode) ~ "backward",
      grepl("_cb$", Mode) ~ "combined"
    )
  ) %>%
  group_by(model, Mode, M, k_thres) %>%
  summarise(rel_nrefits = round(mean(rel_nrefits), 2)) %>%
  ungroup() %>%
  spread("model", "rel_nrefits") %>%
  group_by(Mode) %>%
  mutate(M = ifelse(duplicated(M), "", M)) %>%
  ungroup() %>%
  mutate(Mode = ifelse(duplicated(Mode), "", Mode)) %>%
  rename(`$\\tau$` = "k_thres") %>%
  kable(
    caption = "Mean proportions of required refits for both forward and backward PSIS-LFO-CV.",
    booktabs = TRUE,
    escape = FALSE
  ) %>%
  footnote(
    general = "Note: Results are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. Abbreviations: $\\\\tau$ = threshold of the Pareto $k$ estimates; $M$ = number of predicted future observations.",
    general_title = "",
    threeparttable = TRUE,
    escape = FALSE
  )
```

We may even combine forward and backward mode PSIS-LFO-CV in the following way.
First, we start with forward mode until a refit becomes necessary, say at
observation $i^\star$. Then, we apply backward mode on the basis of the refitted
model and perform multiple proposal importance sampling [@veach1995; @he2014] to
obtain the ELPD values of the observations $i^\star - 1, i^\star - 2, \ldots$
from the mixture of the forward and backward distributions. We do this until the
backward mode requires a refit at which point we stop the process and continue
with forward mode at observation $i^\star$. This algorithm requires exactly as
many refits as the forward mode while potentially increasing accuracy for those
observations for which the pointwise ELPD contribution was computed via both
forward and backward mode PSIS-LFO-CV. In the present paper, we did not
investigate the possibility of multiple importance sampling in more detail, but
it could be a promising extention to be studied in the future.

## Appendix C: PSIS-LFO-CV for the RMSE {-}

We may also use other measures of predictive performance than the ELPD,
for instance the RMSE. For a scalar response $y$ and corresponding vector 
$\hat{y}$ of a toal of $S$ posterior predictions $\hat{y}^{(s)}$, the RSME 
is defined as 

\begin{equation}
\text{RMSE}(y, \hat{y}) = \frac{1}{S} \sum_{s=1}^S (\hat{y}^{(s)} - y)^2.
\end{equation}

If we predict multiple responses in the future (i.e., perform $M$-SAP with
$M > 1$), we simply sum the RMSE over all those responses.
When approximating the RMSE via PSIS, we use the (Pareto smoothed)
importance weights $w^{(s)}$ (see Section \ref{#psis-MSAP}) to estimate

\begin{equation}
\text{RMSE}(y, \hat{y}) \approx
  \frac{\sum_{s=1}^S w^{(s)} (\hat{y}^{(s)} - y)^2}{\sum_{s=1}^S w^{(s)}}.
\end{equation}

The remaining computations are analogous to using the ELPD as a measure of
predictive performance in LFO-CV and so we do not spell out the details here.
The code we provide on GitHub (https://github.com/paul-buerkner/LFO-CV-paper) is
modularized and also has an implementation of the (approximate) RMSE for LFO-CV.

Results of the 1-SAP and 4-SAP RMSE simulations are visualized in Figure
\@ref(fig:1sap-rmse) and \@ref(fig:4sap-rmse), respectively. It is clearly
visible that the the accuracy of the PSIS RMSE approximation is nearly
independent of the threshold $\tau$ when $\tau$ is within the interval
$[0.5,0.7]$ motivated in \@ref(psis) [this would not be the case if $\tau$ was
allowed to be larger; @vehtari2019psis]. For all conditions, the PSIS-LFO-CV
approximation is highly accurate, that is, both approximately unbiased and low
in variance around the corresponding exact LFO-CV RMSE value (represented by the
dashed line in Figure \@ref(fig:1sap)). Taken together, these simulations
indicate that PSIS-LFO-CV not only works well with the ELPD but also with the
RMSE.

```{r}
# data preparation
lfo_sims_rmse <- read_rds("results/lfo_sims_rmse.rds") %>%
  as_tibble() %>%
  filter(lengths(res) > 0) %>%
  mutate(
    model = factor(model, levels = mlevels),
    tau = factor(k_thres, labels = tau_levels),
    # compute sum of pointwise ELPD values
    rmse_exact_lfo = map_dbl(res, ~ get_elpd(.$lfo_exact_rmses)),
    rmse_approx_fw_lfo = map_dbl(res, ~ get_elpd(.$lfo_approx_fw_rmses)),
    # compute differences to the exact values
    rmse_diff_fw_lfo = rmse_approx_fw_lfo - rmse_exact_lfo,
    nrefits_fw = map_dbl(res, ~ get_refits(.$lfo_approx_fw_rmse)),
    rel_nrefits_fw = nrefits_fw / (N - L)
  ) %>%
  select(-res)
```

```{r 1sap-rmse, fig.height=8, fig.cap="Simulation results of 1-step-ahead predictions using the RMSE as measure of predictive accuracy. Histograms are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. The black dashed lines indicates the exact LFO-CV result."}
lfo_sims_rmse %>% 
  filter(is.na(B), M == 1) %>% 
  select(rmse_diff_fw_lfo, model, tau) %>%
  ggplot(aes(x = rmse_diff_fw_lfo, y = ..density..)) +
  facet_grid(
    model ~ tau, scales = "free_y",
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7, fill = elpd_colors[c("lfo_fw")], bins = 20) +
  labs(x = "RMSE difference of approximate and exact LFO-CV", y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    legend.position = "bottom"
  ) +
  NULL
```

```{r 4sap-rmse, fig.height=8, fig.cap="Simulation results of 4-step-ahead predictions using the RMSE as measure of predictive accuracy. Histograms are based on 100 simulation trials of time series with $N = 200$ observations requiring at least $L = 25$ observations to make predictions. The black dashed lines indicates the exact LFO-CV result."}
lfo_sims_rmse %>% 
  filter(is.na(B), M == 4) %>% 
  select(rmse_diff_fw_lfo, model, tau) %>%
  ggplot(aes(x = rmse_diff_fw_lfo, y = ..density..)) +
  facet_grid(
    model ~ tau, scales = "free_y",
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7, fill = elpd_colors[c("lfo_fw")], bins = 20) +
  labs(x = "RMSE difference of approximate and exact LFO-CV", y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  theme_bw() +
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    legend.position = "bottom"
  ) +
  NULL
```

