---
title: "Approximate leave-future-out cross-validation for Bayesian time series models"
author: "Paul Bürkner, Jonah Gabry, Aki Vehtari"
# date: "`r Sys.Date()`"
output: 
 beamer_presentation:
   #toc: True
   theme: "metropolis"
   fig_width: 4  # 6
   fig_height: 2.8  # 4.5
   df_print: kable
   pandoc_args: "--pdf-engine=xelatex"
   slide_level: 2
   includes:
      in_header: theme_options.tex
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, cache=FALSE}
options(width = 100)
knitr::opts_chunk$set(
   cache = TRUE,
   echo = FALSE
)

# define a hook to allow font size changes in beamer:
# from https://stackoverflow.com/questions/26372138/beamer-presentation-rstudio-change-font-size-for-chunk
knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})

library(tidyverse)
library(brms)
library(latex2exp)
theme_set(theme_default())
colors <- unname(unlist(bayesplot::color_scheme_get()[c(6, 2)]))
source("sim_functions.R")
```

## Overarching Goal

\centering
\Large Estimate out-of-sample predictive performance of Bayesian models with high efficiency

## Water Level of Lake Huron

```{r}
data("LakeHuron")
N <- length(LakeHuron)
df <- data.frame(
  y = as.numeric(LakeHuron),
  year = as.numeric(time(LakeHuron)),
  time = 1:N
) 
```

```{r fit_lh}
fit_lh <- brm(
  y | mi() ~ 1, 
  data = df, 
  autocor = cor_ar(~time, p = 4), 
  prior = prior(normal(0, 0.5), class = "ar"),
  chains = 2, warmup = 1000, iter = 5000,
  control = list(adapt_delta = 0.99),
  seed = 5838296, file = "models/fit_lh"
)
```

```{r lake-huron-pred1, cache=FALSE}
oos <- 92:98
df$oos <- factor(seq_len(nrow(df)) %in% oos)
df_sub <- df[-oos, ]
preds <- posterior_predict(fit_lh, newdata = df_sub)
preds <- cbind(
  Estimate = colMeans(preds), 
  Q5 = apply(preds, 2, quantile, probs = 0.05),
  Q95 = apply(preds, 2, quantile, probs = 0.95)
)
preds <- cbind(df_sub, preds)

ggplot(preds, aes(x = year, y = Estimate)) +
  geom_smooth(
    aes(ymin = Q5, ymax = Q95), stat = "identity",
    size = 0.5, fill = "grey60") +
  geom_point(aes(y = y)) + 
  labs(y = "Water Level (ft)", x = "Year") +
  xlim(c(1875, 1975))
```

## Water Level of Lake Huron: Predictions

```{r lake-huron-pred2, cache=FALSE}
preds <- posterior_predict(fit_lh, oos = oos)
preds <- cbind(
  Estimate = colMeans(preds), 
  Q5 = apply(preds, 2, quantile, probs = 0.05),
  Q95 = apply(preds, 2, quantile, probs = 0.95)
)
preds <- cbind(df, preds)
   
ggplot(preds, aes(x = year, y = Estimate)) +
  geom_smooth(
     aes(ymin = Q5, ymax = Q95, fill = oos), 
     stat = "identity", size = 0.5
  ) +
  geom_point(aes(y = y)) + 
  labs(y = "Water Level (ft)", x = "Year") +
  xlim(c(1875, 1975)) +
  scale_fill_manual(values = c("grey60", "red")) +
  guides(fill = FALSE) 
```

## Leave-Future-Out Cross-Validation (LFO-CV)

Perform M-step-ahead predictions (M-SAP) at observation $i$
$$
p(y_{i + 1}, \ldots, y_{i + M} \,|\, y_{1},...,y_{i}) =: 
p(y_{i+1:M} \,|\, y_{1:i})
$$

Estimate expected M-SAP performance via LFO-CV
$$
{\rm ELPD}_{\rm LFO} = \sum_{i=L}^{N - M} \log p(y_{i+1:M} \,|\, y_{1:i})
$$

This requires fitting a separate model for each $i$
$$
p(y_{i+1:M} \,| \, y_{1:i}) = 
  \int p(y_{i+1:M} \,| \, y_{1:i}, \theta) \, 
    p(\theta\,|\,y_{1:i}) \, {\rm d} \theta
$$

<!--
- $N$: total number of observations
- $M$: number of predicted future values 
- $L$: minimal number of observation required for predictions
-->

## Approximate M-Step-Ahead Predictions

\begin{center}
We are moving \textbf{forward} in time!
\end{center}

```{r vis-msap}
status_levels <- c("included", "included (PSIS)", "left out")
df <- data.frame(
  obs = rep(1:9, 3),
  i = factor(rep(3:5, each = 9)),
  Status = c(
    rep("included", 4), rep("left out", 5),
    rep("included", 4), rep("included (PSIS)", 1), rep("left out", 4),
    rep("included", 4), rep("included (PSIS)", 2), rep("left out", 3)
  )
) %>%
  mutate(Status = factor(Status, levels = status_levels))

msap_colors <- c(
  bayesplot::color_scheme_get("viridis")$light,
  bayesplot::color_scheme_get("viridis")$light_highlight,
  bayesplot::color_scheme_get("viridis")$dark
)

ggplot(df, aes(obs, i, fill = Status)) +
  geom_tile(height = 0.9, width = 1, col = "black") +
  annotate(
    'text', x = 5:7, y = c(1, 2, 3), 
    label = "X", parse = TRUE, 
    size = 10, color = "white"
  ) +
  labs(x = "Observation", y = "Predicted observation") +
  scale_x_continuous(breaks = 1:9) +
  scale_fill_manual(values = msap_colors) +
  bayesplot::theme_default() +
  guides(fill = FALSE)
  NULL
```

## Pareto Smoothg Importance Sampling (PSIS) for LFO-CV

PSIS approximation of M-SAP:
$$
 p(y_{i+1:M} \,|\, y_{1:i}) \approx
   \frac{ \sum_{s=1}^S w_i^{(s)} \, p(y_{i+1:M} \,|\, y_{1:i}, \theta^{(s)})}
        { \sum_{s=1}^S w_i^{(s)}}
$$

\bigskip

Let's call $J_i$ the index set of observations included 
in the target model but **not** in the approximating model

For observation $i$ and posterior sample $s$ we compute the importance ratio as
$$
r_i^{(s)} = \prod_{j \in J_i} p(y_j \,|\, \,\theta^{(s)})
$$

Stabilize $r_i^{(s)}$ via Pareto smoothing to obtain weights $w_i^{(s)}$


##

\centering
\Large At what point do we have to refit the model?

## The Generalized Pareto Distribution

```{r}
x <- seq(0.01, 10, 0.01)
k <- c(0, 0.5, 0.7, 1)
y <- unlist(lapply(k, SpatialExtremes::dgpd, x = x, loc = 0, scale = 1))
gpd_df <- data.frame(x, y) %>%
  mutate(k = factor(rep(k, each = n() / length(k))))

ggplot(gpd_df, aes(x, y, color = k)) +
  geom_line(size = 0.8) +
  labs(x = "", y = "Density")
```

\centering
Refit the model if $k$ exceeds a given threshold $\tau$

## Simulation Conditions

```{r, include = FALSE}
seed <- 1234
set.seed(seed)
N <- 200
time <- seq_len(N)
stime <- scale_unit_interval(time)
models <- c("AR2-only", "AR2-linear", "AR2-quadratic")
AR2_labels <- c("stationary", "linear", "quadratic")

fits <- preds <- setNames(vector("list", length(models)), models)
for (m in names(fits)) {
   file <- paste0("models/fit_", m)
   fits[[m]] <- fit_model(model = m, N = N, seed = seed, file = file)
   pred <- posterior_predict(fits[[m]])
   preds[[m]] <- fits[[m]]$data %>% 
     mutate(       
       time = time, 
       stime = stime,
       Estimate = colMeans(pred), 
       Q5 = apply(pred, 2, quantile, probs = 0.05),
       Q95 = apply(pred, 2, quantile, probs = 0.95),
       model = m
    )
}
preds <- as_tibble(bind_rows(preds)) %>%
   mutate(model = factor(model, levels = models, labels = AR2_labels)) %>%
   select(-`I(stime^2)`)
```

```{r simmodels, fig.height=3.5, fig.width=5}
ggplot(preds, aes(x = time, y = Estimate)) +
  facet_wrap(~model) +
  geom_vline(xintercept = 25, linetype = 3) +
  geom_smooth(aes(ymin = Q5, ymax = Q95), stat = "identity", size = 0.5) +
  geom_point(aes(y = y), size = 0.5) +
  labs(y = "y") 
```


## Simulation Results: ELPD of 1-SAP

```{r}
get_elpd <- function(x) {
  summarize_elpds(x)[1]
}
get_refits <- function(x) {
  sum(attr(x, "refits"), na.rm = TRUE)
}

mlevels <- c(
  "constant", "linear", "quadratic",
  "AR2-only", "AR2-linear", "AR2-quadratic"
)
tau_levels <- TeX(paste0("$\\tau$ = ", c(0.5, 0.6, 0.7)))
lfo_sims <- read_rds("results/lfo_sims_1sap.rds") %>%
  bind_rows(read_rds("results/lfo_sims_4sap.rds")) %>%
  as_tibble() %>%
  mutate(
    model = factor(model, levels = mlevels),
    tau = factor(k_thres, labels = tau_levels),
    elpd_loo = map_dbl(res, ~ .$loo_cv$estimates["elpd_loo", 1]),
    elpd_exact_lfo = map_dbl(res, ~ get_elpd(.$lfo_exact_elpds)),
    elpd_approx_lfo = map_dbl(res, ~ get_elpd(.$lfo_approx_fw_elpds)),
    elpd_diff_lfo = elpd_approx_lfo - elpd_exact_lfo,
    elpd_diff_loo = elpd_loo - elpd_exact_lfo,
    nrefits = map_dbl(res, ~ get_refits(.$lfo_approx_fw_elpds)),
    rel_nrefits = nrefits / (N - L)
  ) %>%
  # only keep results of AR models
  filter(is.na(B), grepl("AR2", model)) %>% 
  mutate(model = factor(model, labels = AR2_labels)) 
```

```{r, cache=FALSE}
lfo_refits_1sap <- lfo_sims %>% 
  filter(M == 1) %>%
  select(model, tau, rel_nrefits) %>%
  group_by(model, tau) %>%
  summarise(rel_nrefits = round(mean(rel_nrefits), 2))
```

```{r 1sap, message=FALSE, warning=FALSE, fig.height=3.5, fig.width=5, cache=FALSE}
lfo_sims %>% 
  filter(M == 1) %>%  
  ggplot(aes(x = elpd_diff_lfo, y = ..density..)) +
  facet_grid(
    model ~ tau, #scales = "free_y", 
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7, fill = colors[1]) +
  labs(x = 'ELPD difference to exact LFO-CV', y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_label(
    aes(x = 2.7, y = 2.3, label = rel_nrefits), 
    data = lfo_refits_1sap, inherit.aes = FALSE
  ) + 
  xlim(c(-2, 4)) + 
  theme_bw() +
  theme(
    legend.position = "bottom",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  NULL 
```

## Simulation Results: ELPD of 4-SAP

```{r cache=FALSE}
lfo_refits_4sap <- lfo_sims %>% 
  filter(M == 4) %>%
  select(model, tau, rel_nrefits) %>%
  group_by(model, tau) %>%
  summarise(rel_nrefits = round(mean(rel_nrefits), 2))
```

```{r 4sap, message=FALSE, warning=FALSE, fig.height=3.5, fig.width=5, cache=FALSE}
lfo_sims %>% 
  filter(M == 4) %>% 
  ggplot(aes(x = elpd_diff_lfo, y = ..density..)) +
  facet_grid(
    model ~ tau, #scales = "free_y", 
    labeller = label_parsed
  ) +
  geom_histogram(alpha = 0.7, fill = colors[1]) +
  labs(x = 'ELPD difference to exact LFO-CV', y = "") +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_label(
    aes(x = 20, y = 0.15, label = rel_nrefits), 
    data = lfo_refits_4sap, inherit.aes = FALSE
  ) + 
  xlim(c(-15, 30)) + 
  theme_bw() +
    theme(
    legend.position = "bottom",
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank()
  ) +
  NULL 
```

## Lake Huron Model: Pareto k Estimates

```{r}
L <- 20
M <- 1
lh_elpd_1sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = M, L = L, 
  file = "results/lh_elpd_1sap_exact.rds"
)
lh_elpd_1sap_approx <- compute_lfo(
  fit_lh, type = "approx", M = M, L = L, 
  file = "results/lh_elpd_1sap_approx_fw.rds"
)
refits <- attributes(lh_elpd_1sap_approx)$refits
nrefits <- length(refits)

sum_lh_elpd_1sap_exact <- summarize_elpds(lh_elpd_1sap_exact)[1]
sum_lh_elpd_1sap_approx <- summarize_elpds(lh_elpd_1sap_approx)[1]
```

```{r}
M <- 4
lh_elpd_4sap_exact <- compute_lfo(
  fit_lh, type = "exact", M = M, L = L, 
  file = "results/lh_elpd_4sap_exact.rds"
)
lh_elpd_4sap_approx <- compute_lfo(
  fit_lh, type = "approx", M = M, L = L, 
  file = "results/lh_elpd_4sap_approx_fw.rds"
)
sum_lh_elpd_4sap_exact <- summarize_elpds(lh_elpd_4sap_exact)[1]
sum_lh_elpd_4sap_approx <- summarize_elpds(lh_elpd_4sap_approx)[1]
```

```{r lh-pareto-k}
N <- length(LakeHuron)
ks <- na.omit(attributes(lh_elpd_1sap_approx)$ks)
ids <- (L + 1):N
plot_ks(ks, ids)
```

## Lake Huron Model: ELPD Estimates

```{r lh-pw-elpd}
lh_pw_elpd <- tibble(
  elpd_exact = na.omit(lh_elpd_1sap_exact),
  elpd_approx = na.omit(lh_elpd_1sap_approx),
  k = na.omit(attributes(lh_elpd_1sap_approx)$ks),
  M = "M = 1"
) %>% bind_rows(
  tibble(
    elpd_exact = na.omit(lh_elpd_4sap_exact),
    elpd_approx = na.omit(lh_elpd_4sap_approx),
    k = na.omit(attributes(lh_elpd_4sap_approx)$ks),
    M = "M = 4"
  )
)
ggplot(lh_pw_elpd, aes(elpd_exact, elpd_approx)) +
  facet_wrap(facets = "M", nrow = 1, ncol = 2, scales = "free") +
  geom_abline(slope = 1) +
  geom_point() +
  labs(y = "Approximate ELPD", x = "Exact ELPD") + 
  theme_bw()
```

## Conclusion

- CV has to respect the model's prediction task
- LFO-CV seems reasonable for time series models
- We can approximate LFO-CV via PSIS
- PSIS-LFO-CV provides a close approximation to exact LFO-CV
- PSIS-LFO-CV improves speed by up to two orders of magnitude

\bigskip

Resources:

- Preprint: https://arxiv.org/abs/1902.06281
- GitHub: https://github.com/paul-buerkner/LFO-CV-paper 
- Email: paul.buerkner@gmail.com


## Importance Sampling

All we care about are expectations (over $f$):
$$
\mathbb{E}_f[h(\theta)] = \int h(\theta) f(\theta) \,d\, \theta
$$

Switch the distribution (from $f$ to $g$) over which to integrate:
$$
\mathbb{E}_f[h(\theta)] = 
  \frac{\int h(\theta) r(\theta) g(\theta) \,d\, \theta}
    {\int r(\theta) g(\theta) \,d\, \theta}
$$

with importance ratios
$$
r(\theta) = \frac{f(\theta)}{g(\theta)}
$$

## Pareto Smoothed Importance Sampling (PSIS)

Suppose we can obtain samples $\theta^{(s)}$ from $g$ and compute importance ratios $r(\theta^{(s)}) =: r^{(s)}$. Then we can approximate 
$$
\mathbb{E}_f[h(\theta)] \approx 
\frac{\sum_{s=1}^S r^{(s)} \, h(\theta^{(s)})}{\sum_{s=1}^S r^{(s)}}
$$

Problem: The importance ratios $r^{(s)}$ tend to be highly unstable

Solution: Stabilize $r^{(s)}$ by applying Pareto Smoothing

- PSIS weights $w^{(s)}$ that replace $r^{(s)}$
- Diagnose accuracy via the Pareto shape parameter $k$

